---
title: "hw-ch7"
author: "Austin Melendez"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Section 7.9: Exercises 3, 4, 5

```{r}
library(ISLR2)
```

## Exercise 3

We are given:
$b_1(X) = X$
$b_2(X) = (X-1)^2 I(X\ge1)$

$\hat{\beta}_0 = 1$
$\hat{\beta}_1 = 1$
$\hat{\beta}_2 = -2$

Estimate curve between $X = -2$ and $X = 2$.

Our Model:
$\hat{Y} = 1+[1\times X]-[2\times (X-1)^2 I(X\ge1)] + \epsilon$

Case 1 $X < 1$:
$$ \hat{Y} = 1 + X $$

Case 2 $X \ge 1$:
$$ \hat{Y} = 1 + X - 2(X-1)^2 $$
Image summary:
* From **−2** to **1** linear line increasing
* Smooth transition at **1**
* After **1** Concave-down parabola

```{r}
x <- seq(-2, 2, length = 200)

y <- ifelse(x < 1,
            1 + x,
            1 + x - 2*(x - 1)^2)

plot(x, y, type = "l", lwd = 2,
     xlab = "X", ylab = "Y")
abline(v = 1, lty = 2)
```

## Exercise 4

We are given:
$b_1(X) = I(0\le X\le 2)-(X-1) I(1\le X\le 2)$
$b_2(X) = (X-3) I(3\le X\le 4)+I(4<X<5)$

$\hat{\beta}_0 = 1$
$\hat{\beta}_1 = 1$
$\hat{\beta}_2 = 3$

Estimate curve between $X = -2$ and $X = 6$.

Model:
$\hat{Y} = 1+[1\times I(0\le X\le 2)-(X-1) I(1\le X\le 2)]+[3\times (X-3) I(3\le X\le 4)+I(4<X<5)] + \epsilon$

Case 1 $X < 0$:
$$ \hat{Y} = 1$$

Case 2 $0 \le X < 1$:
$$ \hat{Y} = 2 $$

Case 3 $1 \le X < 2$:
$$ \hat{Y} = 3 - X $$

Case 4 $2 < X < 3$:
$$ \hat{Y} = 1 $$

Case 5 $3 \le X \le 4$:
$$ \hat{Y} = 3X - 8 $$

Case 6 $4 < X \le 5$:
$$ \hat{Y} = 4 $$

Case 7 $X > 5$:
$$ \hat{Y} = 1 $$

Image summary:
* Flat at **1** (left tail)
* Jump to **2** on [0,1)
* Decreasing line to **1** on [1,2]
* Flat again at **1**
* Sharp increase on [3,4]
* Flat at **4** on (4,5]
* Drop back to **1**

```{r}
x <- seq(-2, 6, length = 400)

y <- ifelse(x < 0, 1,
     ifelse(x < 1, 2,
     ifelse(x <= 2, 3 - x,
     ifelse(x < 3, 1,
     ifelse(x <= 4, 3*x - 8,
     ifelse(x <= 5, 4, 1))))))

plot(x, y, type = "l", lwd = 2,
     xlab = "X", ylab = "Y")

abline(v = c(0,1,2,3,4,5), lty = 3)
```

## Exercise 5
Consider two curves, $\hat{g}_1$ and $\hat{g}_2$, defined by:

$$ \hat{g}_1 = arg\space min(\sum^{n}_{i=1}(y_i - g(x_i))^2 + \lambda \int[g^{(3)}(x)]^2 \space dx)$$

$$ \hat{g}_2 = arg\space min(\sum^{n}_{i=1}(y_i - g(x_i))^2 + \lambda \int[g^{(4)}(x)]^2 \space dx)$$

where $g^{(m)}$ represents the mth derivative of g.

#### A 
As $\lambda \to \infty$, will $\hat{g}_1$ or $\hat{g}_2$ have the smaller training RSS?

$\hat{g}_1$ penalizes 3rd derivative → quadratic (degree 2)
$\hat{g}_2$ penalizes 4th derivative → cubic (degree 3)

Cubic is more flexible and fits training data better. So $\hat{g}_2$ has smaller training RSS. 

#### B 
As $\lambda \to \infty$, will $\hat{g}_1$ or $\hat{g}_2$ have the smaller test RSS?

Cubic has lower bias than quadratic and typically generalizes better. So $\hat{g}_2$ has smaller test RSS. 

#### C
For $\lambda = 0$, will $\hat{g}_1$ or $\hat{g}_2$ have the smaller training and test RSS?

Training RSS = 0 for both meaning they both overfit, which results in high test RSS