---
title: "project3"
author: "Austin Melendez"
date: "2026-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
#install.packages("keras")
#install.packages("magick")
#install.packages("ISLR2")
#install.packages("torch")
#install.packages("torchvision")
#install.packages("tidyverse")
#install.packages("corrplot")

library(keras)
library(magick)
library(ISLR2)
library(torch)
library(torchvision)
library(tidyverse)
library(corrplot)
```

# Project 3: Neural Networks & Clustering

## Part 1: Neural Networks
From your collection of personal photographs, pick 10 images of animals (such as dogs, cats, birds, farm animals, etc.). If the subject does not occupy a reasonable part of the image, then crop the image.

Now use a pretrained image classification CNN as in Lab 10.9.4 to predict the class of each of your images, and report the probabilities for the top five predicted classes for each image.

Make sure to include your original images as part of your project submission. 

### (From Your Revised Textbook Lab 10) 
### Using Pretrained CNN Models
```{r}
setwd("C:/Users/Austin/Desktop/Everything/Coding/sacramento-state-university/STAT140B - Statistical Learning/Assignments/project3data/") # you will need to update this! #
img_dir <- "cnn"
image_names <- list.files(img_dir)
num_images <- length(image_names)
x <- torch_empty(num_images, 3, 224, 224)
for (i in 1:num_images) {
   img_path <- file.path(img_dir, image_names[i])
   img <- img_path %>% 
     base_loader() %>% 
     transform_to_tensor() %>% 
     transform_resize(c(224, 224)) %>% 
     
     # normalize with imagenet mean and stds.
     transform_normalize(
       mean = c(0.485, 0.456, 0.406),
       std = c(0.229, 0.224, 0.225)
     )
   x[i,,, ] <- img
}
```

We then load the trained network. The model has 18 layers, with a fair bit of complexity.

```{r}
model <- torchvision::model_resnet18(pretrained = TRUE)
model$eval() # put the model in evaluation mode
```

Finally, we classify our images, and return the top five class
choices in terms of predicted probability for each.

```{r}
preds <- model(x)

mapping <- jsonlite::read_json("https://s3.amazonaws.com/deep-learning-models/image-models/imagenet_class_index.json") %>% 
  sapply(function(x) x[[2]])

top5 <- torch_topk(preds, dim = 2, k = 5)

top5_prob <- top5[[1]] %>% 
  nnf_softmax(dim = 2) %>% 
  torch_unbind() %>% 
  lapply(as.numeric)

top5_class <- top5[[2]] %>% 
  torch_unbind() %>% 
  lapply(function(x) mapping[as.integer(x)])

result <- purrr::map2(top5_prob, top5_class, function(pr, cl) {
  names(pr) <- cl
  pr
})
names(result) <- image_names
print(result)
```

The images I used are:

```{r}
library(jpeg)
library(png)

img_dir <- "project3data/cnn"

image_names <- c(
  "australian_shepard.jpg",
  "australian_shepard2.jpg",
  "bald_eagle.jpg",
  "banana_slug.jpg",
  "bird.jpg",
  "black_laborador.jpg",
  "deer.jpg",
  "duck.jpg",
  "hawaiian_green_sea_turtle.jpg",
  "yellow_field_laborador.jpg",
  "yellow_field_laborador2.jpg"
)

par(mfrow = c(2,3), mar = c(1,1,2,1))

for(i in 1:length(image_names)){

  img_path <- file.path(img_dir, image_names[i])
  img <- jpeg::readJPEG(img_path)

  plot(1:2,
       type = "n",
       axes = FALSE,
       xlab = "",
       ylab = "",
       main = image_names[i])

  rasterImage(img, 1, 1, 2, 2)
}
```

For further reference, the full-sized images I used are uploaded and labeled here on [Imgur.com](https://imgur.com/a/DraPV2s):

## Part 2: Unsupervised Learning

The dataset I used was [The Repeat Consumption Matrices dataset](https://archive.ics.uci.edu/dataset/441/repeat+consumption+matrices). It contains consumer repeat-purchase behavior data, each matrix contains how many times a user 'consumed' and item. Items can be locations, artists, or subreddits.

The goal of this analysis is to identify patterns in repeat consumer purchasing behavior using several unsupervised learning techniques discussed in ISLR2. Specifically, this report applies:

- Principal Components Analysis (PCA)
- K-means clustering
- Hierarchical clustering

## 1. Dataset Description

The Repeat Consumption Matrices dataset contains repeat-purchase behavior information for consumers. The data measure patterns of repeated item consumption and are useful for identifying groups of consumers with similar purchasing behavior.

The dataset contains user-item consumption matrices from several online platforms such as Reddit, Twitter, Gowalla, and LastFM. Each row represents a user and each column represents an item. Entries represent the number of times a user consumed that item.

The **primary goal** is to determine whether users naturally separate into distinct groups based on their listening behavior.

## 2. Data Description

According to the data dictionary:

**lastfm** *(992 x 15k)*: How many times a user listened to each artist.

So the LastFM subset contains:
- 992 users
- ~15,000 artists

## 3. Loading Data

```{r}
setwd("C:/Users/Austin/Desktop/Everything/Coding/sacramento-state-university/STAT140B - Statistical Learning/Assignments/")

data_dir <- "project3data"
url <- "lastfm/train.csv"

raw_data <- read.csv(file.path(data_dir, url), header = FALSE)

colnames(raw_data) <- c("user", "artist", "count")

top_users <- raw_data %>%
  group_by(user) %>%
  summarise(total = sum(count)) %>%
  arrange(desc(total)) %>%
  slice(1:150)

top_artists <- raw_data %>%
  group_by(artist) %>%
  summarise(total = sum(count)) %>%
  arrange(desc(total)) %>%
  slice(1:100)

filtered <- raw_data %>%
  filter(user %in% top_users$user,
         artist %in% top_artists$artist)

music_matrix <- xtabs(count ~ user + artist, data = filtered)

dim(music_matrix)
```

## 4. Exploratory Data Analysis

```{r}
counts <- as.vector(music_matrix)

hist(counts,
     breaks = 40,
     main = "Distribution of Listening Counts",
     xlab = "Listening Count")
```

```{r}
counts <- as.vector(music_matrix)

hist(counts[counts > 0 & counts < 100],
     breaks = 50,
     main = "Zoomed Distribution",
     xlab = "Count")
```

```{r}
artist_variance <- apply(music_matrix, 2, var)

hist(artist_variance,
     breaks = 30,
     main = "Variance Across Artists",
     xlab = "Variance")
```

```{r}
top_var_cols <- order(artist_variance, decreasing = TRUE)[1:12]

cor_matrix <- cor(music_matrix[, top_var_cols],
                  use = "pairwise.complete.obs")

corrplot::corrplot(cor_matrix,
                   method = "color",
                   tl.cex = 0.7)
```

## 5. Standardization

```{r}
# remove zero-variance columns
keep <- apply(music_matrix, 2, var) > 0
music_matrix <- music_matrix[, keep]

# convert to plain matrix (important for stability)
music_matrix <- as.matrix(music_matrix)

# scale safely
scaled_matrix <- scale(music_matrix)

# remove any rows with NA/Inf after scaling
scaled_matrix <- scaled_matrix[complete.cases(scaled_matrix), , drop = FALSE]

# remove any remaining degenerate rows (all constant)
scaled_matrix <- scaled_matrix[apply(scaled_matrix, 1, var) > 0, , drop = FALSE]
```

## 6. PCA

```{r}
pca_fit <- prcomp(scaled_matrix)

pve <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2)

plot(pve,
     type = "b",
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot")

scores <- pca_fit$x

plot(scores[,1],
     scores[,2],
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA Score Plot")
```

## 7. K-Means Clustering

```{r}
set.seed(1)

# cap k so it can never exceed number of observations - 1
max_k <- min(10, n - 1)

wss <- numeric(max_k)

for(k in 1:max_k){
  km <- kmeans(scaled_matrix, centers = k, nstart = 20)
  wss[k] <- km$tot.withinss
}

plot(1:max_k, wss,
     type = "b",
     xlab = "Number of Clusters",
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Plot for K-Means Clustering")
```

```{r}
km_fit <- kmeans(scaled_matrix, centers = 3, nstart = 50)

table(km_fit$cluster)

plot(scores[,1],
     scores[,2],
     col = km_fit$cluster,
     pch = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "K-Means Clusters on PCA Scores")
```

## 8. Hierarchical Clustering

```{r}
dist_matrix <- dist(scaled_matrix)

hc_complete <- hclust(dist_matrix, method = "complete")

plot(hc_complete, labels = FALSE)

hc_clusters <- cutree(hc_complete, k = 3)

table(hc_clusters)

plot(scores[,1],
     scores[,2],
     col = hc_clusters,
     pch = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "Hierarchical Clusters on PCA Scores")
```

## Conclusion

PCA and clustering reveal approximately three stable groups of users with consistent structure across methods.
```