---
title: "Example 6.2"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Leverage

```{r}
library(faraway)
data(savings)
mod1 <- lm(sr ~ pop15 + pop75 + dpi + ddpi, savings)
hatv <- hatvalues(mod1) # leverages
sum(hatv)
```
Sum of leverages = number of parameters (4 + 1 intercept)

### Half-normal plots

1. Sort the data: $x_{[1]} \le x_{[2]} \le \dots \le x_{[n]}$
2. Compute $u_i = \Phi^{-1}\left(\frac{n+i}{2n+1}\right)$
3. Plot $x_{[i]}$ against $u_i$

and we look for outliers

```{r}
countries <- row.names(savings)
halfnorm(hatv,labs=countries,ylab="Leverages")
```

### Standardized Residuals

```{r}
qqnorm(rstandard(mod1))
abline(0,1)
# OR #
plot(mod1, 2)
```

## Outliers

```{r}
set.seed(0)
testdata <- data.frame(x=1:10, y=1:10+rnorm(10))
mod2 <- lm(y ~ x, testdata)
plot(testdata)
```

Here, we add an outlier with low leverage/influence

```{r}
p1 <- c(5.5, 12)
mod3 <- lm(y ~ x, rbind(testdata, p1))

plot(y ~ x, rbind(testdata, p1))
points(5.5,12,pch=4,cex=2)
abline(mod2) # original model
abline(mod3, lty=2) # model with outlier
```

Next, a point well outside the range of X (high leverage)

```{r}
p2 <- c(15,15.1)
mod4 <- lm(y ~ x, rbind(testdata, p2))
plot(y ~ x, rbind(testdata, p2))
points(15,15.1,pch=4,cex=2)
abline(mod2)
abline(mod4,lty=2)
```

This point is not an outlier, though, and is not influential. 

Finally, a high leverage outlier that is an influential point

```{r}
p3 <- c(15,5.1)
mod5 <- lm(y ~ x, rbind(testdata, p3))
plot(y ~ x, rbind(testdata, p3))
points(15,5.1,pch=4,cex=2)
abline(mod2)
abline(mod5,lty=2)
```

This time the point has changed the line significantly, and so this point is also influential. 

### Studentized Residuals

```{r}
stud <- rstudent(mod1) # studentized residuals
stud[which.max(abs(stud))]
# unadjusted CV:
qt(.05/2,44)
# Bonferroni-corrected CV:
qt(.05/(50*2),44)
```

Here, $n=50$ and we want a two-sided test, hence $50*2$.

Since $2.85 < 3.53$, using the Bonferonni correction we are unable to conclude that Zambia is an outlier. 

### Example Dataset

```{r}
data(star, package="faraway")
plot(star$temp, star$light, xlab="log(temperature)", ylab="log(light intensity")
```

The four stars in the upper left corner do not match the pattern (positive, linear) seen in the rest of the data.

```{r}
lmod <- lm(light ~ temp, star)

plot(star$temp, star$light, xlab="log(temperature)", ylab="log(light intensity")
abline(lmod)
```

```{r}
range(rstudent(lmod))
```

No outliers, but we clearly have a cluster of weird points. (They're not showing up as outliers precisely *because* they're a cluster.)

It turns out that these stars are gas giants, and perhaps need to be considered separately. Let's try removing them from the model:

```{r}
lmod2 <- lm(light ~ temp, star, subset=temp>3.6)

plot(star$temp, star$light, xlab="log(temperature)", ylab="log(light intensity")
abline(lmod)
abline(lmod2, lty=2)
```

## Influential Observations

```{r}
mod1$call

cook <- cooks.distance(mod1)
halfnorm(cook, 3, labs=countries, ylab="Cook's Distances")
```

We can also use the `plot` command to get a plot with Cook's distances

```{r}
plot(mod1, 4)
```

which identifies the same observations as the halfnormal plot.

Let's try removing the observation with the largest Cook statistic.

```{r}
mod6 <- lm(sr ~ pop15 + pop75 + dpi + ddpi, savings,
            subset = cook < max(cook))
summary(mod6)
```

Which we can compare to the fit with the full data

```{r}
summary(mod1)
```

We see, for example, that `ddpi` coefficient estimate changed by about 50\%, which is a lot for a difference of just one country.

Let's examine the leave-one-out differences for this coefficient.

```{r}
plot(dfbeta(mod1)[,2], ylab="Change in pop15 coef")
abline(h=0)
```

For this next chunk of code, we use the `identify` function to find specific points in a scatter plot. This does not always work in R Markdown, so you may need to copy and paste the entire code chunk into the console.

```{r eval=FALSE}
plot(dfbeta(mod1)[,2], ylab="Change in pop15 coef")
abline(h=0)
identify(dfbeta(mod1)[,2])
```

In the Plots tab, click on the point of interest, then press esc. The location of the point(s) you clicked on will appear as output in the console. 

```{r}
savings[23,]
```

```{r}
modj <- lm(sr ~ pop15+pop75+dpi+ddpi,savings,
           subset=(countries != "Japan"))
summary(modj)
```

Compared to lmod1, the ddpi term is no longer significant. $R^2$ has also decreased quite a bit. 

We can also use the plot function on a linear model without specifying which plot we want. This will give us a bunch of useful plots all at once.

```{r}
par(mfrow=c(2,2))
plot(mod1)
par(mfrow=c(1,1))
```

Personally, for diagnostic plots, I prefer

```{r}
plot(mod1, c(1,2,4))
```