---
title: "Chapter 14 Example"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 16 Exercise 6

```{r}
library(faraway)
data("hsb")

str(hsb)
summary(hsb)
```

### Part A 

Model the `math` score in terms of the five factors: `gender`, `race`, `ses`, `schtyp` and
`prog`. Include all second-order interactions but no higher order interactions.
How many regression parameters does your model use? Explain how this can
be calculated in terms of the number of levels for each factor.

```{r}
mod1 <- lm(math ~ (gender + race + ses + schtyp + prog)^2, hsb)
summary(mod1)
```

Our 5 factors `gender`, `race`, `ses`, `schtyp` and `prog` have $i=2$, $j=4$, $k=3$, $l=2$, and $m=3$ levels, respectively. 
So there are 
\[\begin{aligned}
  &(i-1)+(j-1)+(k-1)+(l-1)+(m-1) \\
  &=(2-1)+(4-1)+(3-1)+(2-1)+(3-1)\\
  &=9
\end{aligned}\]
non-interaction terms in the model.
There then needs to be a parameter for each interaction between the dummy variables for a specific factor and all of the dummy variables for the remaining factors: 
\[\begin{aligned}
  &(i-1)[(j-1)+(k-1)+(l-1)+(m-1)] \\
  &\quad + (j-1)[(k-1)+(l-1)+(m-1)] \\
  &\quad + (k-1)[(l-1)+(m-1)] + (l-1)(m-1) \\
  &= (2-1)[(4-1)+(3-1)+(2-1)+(3-1)] \\
  &\quad + (4-1)[(3-1)+(2-1)+(3-1)] \\
  &\quad + (3-1)[(2-1)+(3-1)] + (2-1)(3-1) \\
  &= 1(3+2+1+2) + 3(2+1+2) + 2(1+2) + 1(2) \\
  &= 31
\end{aligned}\]
plus the interaction term is $9+31+1=41$ regression parameters.

In the R output, we can see this quickly by looking at the model degrees of freedom. 

### Part B

Determine whether some two-way interactions can be eliminated using the
anova function.

We are interacting the following variables:

- `gender:race`
- `gender:ses`
- `gender:schtyp`
- `gender:prog`
- `race:ses`
- `race:schtyp`
- `race:prog`
- `ses:schtyp`
- `ses:prog`
- `schtyp:prog`

I'm going to remove a few at a time to speed things up.

```{r}
mod2 <- update(mod1, . ~ .  -gender:race - gender:ses - gender:schtyp - gender:prog)
anova(mod2, mod1) 

mod3 <- update(mod2, . ~ . -race:ses -race:schtyp -race:prog)
anova(mod3, mod2)

mod4 <- update(mod3, . ~ . -ses:schtyp -ses:prog)
anova(mod4, mod3)

mod5 <- update(mod4, . ~ . -schtyp:prog)
anova(mod5, mod4)
```

### Part C 

Determine whether some two-way interactions can be eliminated but now using
the drop1 function. Why do the results differ from the previous question?
Which method of testing should be preferred?

```{r}
drop1(mod1, test="F")
```

So in this case we would drop everything *except* the `race:prog` interaction. 

This approach drops one at a time, whereas in the previous approach I dropped a few at a time. 

### Part D

Fit a model with only main effects and compare it to the model with all two-way
interactions. Which model should be preferred?

This is my `mod5` versus `mod1`.

```{r}
anova(mod5, mod1)
summary(mod5)
anova(mod5)
```

### Part E

Determine which main effects are statistically significant and report on the
differences you find.

```{r}
mod6 <- update(mod5,  .~. -gender)
anova(mod6)
mod7 <- update(mod6, .~. -schtyp)
anova(mod7)
summary(mod7)

TukeyHSD(aov(mod7))
```

### Part F

Check the regression diagnostics for the main effects only model and report on
any anomalies.

```{r}
plot(mod7, 1)
plot(mod7, 2)
```