---
title: "Multiple Comparisons"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set()
```

## ANOVA diagnostics

```{r}
amod <- aov(weight ~ feed, chickwts)
anova(amod)
```

## Diagnostics

```{r}
par(mfrow=c(1,2))
plot(amod, c(1,2))
```


## After the ANOVA

- $H_0: \mu_1 = \mu_2 = \dots = \mu_k$ 
- $H_A: \mu_i \ne \mu_j$ for at least one pair $(i, j)$
  
  We've rejected the null hypothesis... now what?
  
## Pairwise Comparisons

A confidence interval for the difference between two levels can be formulated as 

$$\hat{\mu}_i-\hat{\mu}_j \pm t_{df}^{\alpha/2}SE(\hat{\mu}_i-\hat{\mu}_j)$$

where $SE(\hat{\mu}_i-\hat{\mu}_j) = \hat{\sigma}\sqrt{1/n_i + 1/n_j}$ and $df=n-k$

## Multiple Comparisons

Why do the ANOVA at all?

Why not just start with pairwise comparisons?

## Type I Error Rate

For the `chickwts` data, there are 6 different feeds.

- Examining all possible pairs means doing 15 different comparisons.
- Even if there are no differences, with an $\alpha=0.05$, there's a pretty good chance one of them will be statistically significant just by chance!
- In a multiple comparisons setting, the true Type I error rate is *higher than* $\alpha$
    - We need to adjust for this
    
## Multiple Comparisons

Note: If we only want to look at a few specific pairwise comparisons, we need to select those *before looking at the data*

- If we only look at the comparisons that look interesting based on the data, we are implicitly making all the other comparisons as well. 

## Test Corrections

- Tukey's HSD
- Bonferroni correction
- Benjamini-Hochberg
- ...and many more

## Tukey's HSD (honest significant difference)

- This method is a good balance of dealing with Type I error without being too aggressive.
- Essentially a t-test, but one that corrects for family-wise error rate (multiple comparisons).

## Tukey's HSD

- Compare absolute t values, $|t| = \left|\frac{\bar{x}_i - \bar{x}_j}{SE}\right|$ to the *studentized range distribution*
  - Adjusted p-value
- Family-wise confidence intervals

## studentized range distribution

Let $X_1, \dots, X_n$ be iid $N(\mu, \sigma^2)$ and let $R$ be the range of $X$. 

Then $q_{n, \nu} = R/\hat{\sigma}$ has the studentized range distribution where $\nu$ is the degrees of freedom used in estimating $\hat{\sigma}$. 

## Tukey CIs

$$\hat{\mu}_i-\hat{\mu}_j \pm \frac{q_{n, \nu}}{\sqrt{2}}\hat{\sigma}\left(\frac{1}{n_i}+\frac{1}{n_j}\right)$$

Note: when the group sizes are very unequal, Tukey's HSD may become too conservative. 

## Example: Chick Weights

```{r}
anova(amod)
```

***

```{r eval=FALSE}
TukeyHSD(amod)
```

```
                           diff         lwr       upr     p adj
horsebean-casein    -163.383333 -232.346876 -94.41979 0.0000000
linseed-casein      -104.833333 -170.587491 -39.07918 0.0002100
meatmeal-casein      -46.674242 -113.906207  20.55772 0.3324584
soybean-casein       -77.154762 -140.517054 -13.79247 0.0083653
sunflower-casein       5.333333  -60.420825  71.08749 0.9998902
linseed-horsebean     58.550000  -10.413543 127.51354 0.1413329
meatmeal-horsebean   116.709091   46.335105 187.08308 0.0001062
soybean-horsebean     86.228571   19.541684 152.91546 0.0042167
sunflower-horsebean  168.716667   99.753124 237.68021 0.0000000
meatmeal-linseed      58.159091   -9.072873 125.39106 0.1276965
soybean-linseed       27.678571  -35.683721  91.04086 0.7932853
sunflower-linseed    110.166667   44.412509 175.92082 0.0000884
soybean-meatmeal     -30.480519  -95.375109  34.41407 0.7391356
sunflower-meatmeal    52.007576  -15.224388 119.23954 0.2206962
sunflower-soybean     82.488095   19.125803 145.85039 0.0038845
```

## Example: Chick Weights

```{r}
plot(TukeyHSD(aov(weight ~ feed, chickwts)))
```

## Example: Junior School Project

Data on mathematics test scores for students in 49 London school.

```{r}
data(jsp, package="faraway")
jsp$mathcent <- jsp$math - mean(jsp$math)
```

We're interested in deviations from the average, so we center the scores.

##  Example: Junior School Project

```{r}
plot(jsp$mathcent ~ jsp$school)
```

##  Example: Junior School Project

```{r echo=FALSE}
lmod <- lm(mathcent ~ school-1, jsp)
```

```{r eval=FALSE}
lmod <- lm(mathcent ~ school-1, jsp) # no-intercept model
summary(lmod)
```

```
Coefficients:
         Estimate Std. Error t value Pr(>|t|)    
school1  -3.36845    0.76861  -4.383 1.21e-05 ***
school2   0.67141    1.22870   0.546 0.584806    
school3   2.29641    1.06409   2.158 0.030995 *  
... 
school49  2.79641    0.86882   3.219 0.001301 ** 
school50 -2.65203    0.73356  -3.615 0.000305 ***
---

Residual standard error: 7.372 on 3187 degrees of freedom
Multiple R-squared:  0.08206,	Adjusted R-squared:  0.06794 
F-statistic: 5.814 on 49 and 3187 DF,  p-value: < 2.2e-16
```

##  Example: Junior School Project

- The no-intercept model allows us to examine all 50 schools
- Since we've centered the response, the t-tests are meaningful
    - Each is testing a difference from 0, the center
- Good initial evidence for differences, but we're doing a lot of multiple comparisons here!

##  Example: Junior School Project

```{r}
anova(lmod)
```

##  Example: Junior School Project

- There are a ton of possible comparisons here! 
- Examine which schools are significantly different from average (rather than pairwise between each other).
- We still need to make an adjustment for multiple comparisons

## Bonferroni Correction

- Controls the *family-wise error rate*
    - the overall probability of falsely detecting a difference if none exists
- Adjust $\alpha$ by dividing by the number of comparisons
    - Or, equivalently, multiply all p-values by the number of comparisons.
    
## Bonferroni Correction

```{r results=FALSE}
pvals <- summary(lmod)$coef[,4]
padj <- p.adjust(pvals, method="bonferroni")
coef(lmod)[padj < 0.05]
```

```
  school1  school16  school21  school28   
-3.368450 -3.737400 -3.618450 -5.828595   

school31  school40  school45  school50
3.824053 -4.985458 -4.639201 -2.652027
```

## Benjamini and Hochberg

- Controlling for the family-wise error rate can be overly conservative.
- The Benjamini and Hochberg correction instead controls for the *false discovery rate*
    - This is the proportion of effects identified as significant which are not real.

## Benjamini and Hochberg

Given sorted p-values $p_{(i)}$ for $i = 1, \dots, m$, find the largest $i$ for which $$p_{(i)} \le \alpha\left(\frac{i}{m}\right)$$All tests corresponding to $p_{(i)}$ up to and including this $i$ are considered significant. 

## Benjamini and Hochberg

For example, suppose $m=5$ and we have sorted p-values $\{0.002, 0.005, 0.035, 0.056, 0.122\}$.

We compare these p-values to $$\alpha\left(\frac{i}{m}\right) = 0.05\frac{i}{5} = 0.01i$$
or the values $\{0.01, 0.02, 0.03, 0.04, 0.05\}$

The largest $i$ for which $p_{(i)} \le \alpha\left(\frac{i}{m}\right)$ is $i=2$, where $0.005 < 0.02$

## Example

For our school data,

```{r}
names(which(sort(pvals) < (1:49)*0.05/49))
```

## Example

Or we can use the code

```{r results=FALSE}
padj <- p.adjust(pvals, method="fdr")
coef(lmod)[padj < 0.05]
```

```
  school1   school4   school9  school14  school16  school21  school24   
-3.368450 -2.661928  2.245764 -3.289835 -3.737400 -3.618450  2.623786 

 school27  school28  school31 school34  school36  school40  school45   
-2.305764 -5.828595  3.824053 1.935898  2.360544 -4.985458 -4.639201  

school46  school47  school49  school50
3.063562  2.396895  2.796405 -2.652027
```

## Example

In this case, 18 schools are identified rather than the 8 identified by the Bonferroni approach.

In general, adjusting for false discovery rate is less stringent than adjusting for family-wise error rate.