---
title: "Example 6.1"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Constant Variance

```{r}
library(faraway)
data(savings)
mod1 <- lm(sr ~., savings)
summary(mod1)
```

### Residuals vs. fitted values

```{r}
par(mfrow=c(1,2))
plot(fitted(mod1), residuals(mod1), xlab="Fitted", ylab="Residuals")
abline(h=0)

plot(fitted(mod1), abs(residuals(mod1)), xlab="Fitted", ylab="|Residuals|")
par(mfrow=c (1, 1))
```

Some sample plots

```{r}
par(mfrow=c(2,2))
# no issues
for (i in 1:4) plot (1:50, rnorm (50)); abline(h=0)
# strong nonconstant variance
for (i in 1:4) plot (1:50, (1:50)*rnorm(50)); abline(h=0)
# mild nonconstant variance
for (i in 1:4) plot (1:50, sqrt ((1:50))*rnorm(50)); abline(h=0)
# nonlinearity
for (i in 1:4) plot(1:50, cos ((1:50)*pi/25)+rnorm(50)); abline(h=0)
par(mfrow=c(1, 1))
```

Looking at sample plots can help us get a better "feel" for actual problems vs. random variation.

### Transforming the response

```{r}
data(gala)
mod2 <- lm(Species ~ Area + Elevation + Scruz + Nearest + Adjacent, gala)
plot(fitted (mod2), residuals(mod2), xlab="Fitted", ylab="Residuals")
abline(h=0)
```

Since these are count data, let's try a square root transformation.

```{r}
mod3 <- lm(sqrt(Species) ~ Area + Elevation + Scruz + Nearest + Adjacent, gala)
plot(fitted(mod3), residuals(mod3), xlab="Fitted", ylab="Residuals")
abline(h=0)
```

### Box-Cox Transformation

```{r}
library(sp)
library(gstat)
data(meuse)
names(meuse)
```

```{r}
mod4 <- lm(zinc ~ elev + dist, data=meuse)
plot(fitted(mod4), residuals(mod4), xlab="Fitted", ylab="Residuals")
abline(h=0)
```

Run the box-cox optimization

```{r}
library(MASS)
bc <- boxcox(mod4)
```

Get lambda

```{r}
str(bc)
lam <- bc$x[which.max(bc$y)]
lam
```

Transformation of y

```{r}
BCTransform <- function(y, lambda=0){
  if(lambda == 0) log(y)
  else ((y^lambda - 1) / lambda)
}

BCTransformInverse <- function(yt, lambda=0){
  if (lambda == 0) exp(yt)
  else exp(log(1 + lambda * yt)/lambda)
}

meuse$yt <- BCTransform(meuse$zinc, lam)
```

New model with transformed outcome

```{r}
mod5 <- lm(yt ~ elev + dist, data=meuse)
summary(mod5)
```

Compare to original plot

```{r}
par(mfrow=c(1,2))
plot(fitted(mod4), residuals(mod4), xlab="Fitted", ylab="Residuals")
abline(h=0)
plot(fitted(mod5), residuals(mod5), xlab="Fitted", ylab="Residuals")
abline(h=0)
par(mfrow=c(1,1))
```

```{r}
plot(mod4, 1)
```

## Normality of Errors

```{r}
qqnorm(residuals(mod1), ylab="Residuals", main="")
qqline(residuals(mod1))
# OR #
plot(mod1, 2)
```

This provides a much more exact means of examining normality than a histogram, which can vary significantly depending on how the bins get set. 

### Examining Problems

```{r}
par(mfrow=c(2,2))
n <- 50
# normal
for(i in 1:4){x <- rnorm(n); qqnorm(x); qqline(x)}
# longormal (skewed)
for(i in 1:4){x <- exp(rnorm(n)); qqnorm(x); qqline(x)}
# cauchy (long-tailed)
for(i in 1:4){x <- rcauchy(n); qqnorm(x); qqline(x)}
# uniform
for(i in 1:4){x <- runif(n); qqnorm(x); qqline(x)}
par(mfrow=c(1,1))
```

Q-Q plots can be hard to use! Sometimes there's obviously a problem, but sometimes it's not clear if we have long-tailed error or just outliers. We can check by removing potential outliers and rerunning the model.

