---
title: "Example 7.2"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rescaling

```{r}
data(savings, package="faraway")
summary(savings)
```

`dpi` is on a much different scale than the other variables. Let's change it to thousands of dollars.

```{r}
lmod <- lm(sr ~ pop15+pop75+dpi+ddpi,savings)
summary(lmod)

lmod2 <- lm(sr ~ pop15+pop75+I(dpi/1000)+ddpi,savings)
summary(lmod2)
```

The only change is the rescaling of our coefficient estimate and its associated standard error.

Sometimes we may want to standardize all of our variables to get everything on exactly the same scale.
Typically we convert to mean 0, standard deviation 1.

```{r}
head(savings)
scsav <- data.frame(scale(savings)) 
lmod3 <- lm(sr ~ ., scsav)
summary(lmod3)

scsav2 <- data.frame(sr=savings$sr, scale(savings[,-1]))
lmod4 <- lm(sr ~ ., scsav2)
summary(lmod4)
```

Rescaling everything forces the intercept to 0; rescaling only all input variables does not.