---
title: "Block Designs"
author: "Prof. Lauren Perry"
output:
  ioslides_presentation: default
  beamer_presentation: default
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(faraway)
```

## Completely Randomized Designs

In a completely randomized design (CRD) treatments are assigned to experimental units *at random*.

- Appropriate when units are homogeneous, but what if they're not?
- If we don't know how to describe the form of the heterogeneity, we can continue to use CRD.
- When units are heterogeneous in a known way, we can use *blocks*.

## Blocks

- Idea: experimental units are heterogeneous in a known way and can be arranged
into blocks.
    - within-block variation ideally small
    - between block variation ideally large
    - a block design can then be more efficient than a CRD
- We prefer to have a block size equal to the number of treatments
    - If this cannot be done, an incomplete block design must be used

## Blocks

- Blocks may be determined by the experimenter
  - Ex: we think the treatment will function differently for high- vs low-risk patients, so we use risk level as a block
- Blocks may also be predetermined by the nature of the experiment
  - Ex: a study on crops in three different fields, where we use fields as a block

## Randomized Block Designs

- In a randomized block design, the treatment levels are assigned randomly within a
block.
    - Randomization restricted relative to CRD
    - This means fewer possible permutations for random treatment assignment
- Because they are a feature of the experimental, block effects must be included in the model regardless of significance
    - That is, we cannot regain these degrees of freedom if not significant

## Randomized Block Designs

Suppose we have one treatment factor $\tau$ at $t$ levels and one blocking facor $\beta$ at $r$ levels: $$y_{ij}=\mu + \tau_i + \beta_j + \epsilon_{ij}$$

- Suppose there is one observation for each treatment in each block. 
- This is a *randomized complete block design*. 
- Analysis very similar to two-way ANOVA
    - Model can be extended to include interactions. 
    - Significance of block effect only useful for future reference. 
    
## Example

```{r}
data(oatvar)
attach(oatvar)
head(oatvar)
```

## Example

```{r}
xtabs(yield ~ variety + block)
```

## Example

```{r}
par(mfrow=c(1,2))
boxplot(yield ~ variety, xlab="yield", ylab="variety")
boxplot(yield ~ block, xlab="yield",ylab="block")
```

## Example

```{r}
par(mfrow=c(1,2))
interaction.plot(variety, block, yield)
interaction.plot(block, variety, yield)
```

## Example

```{r}
mod1 <- lm(yield ~ block+variety)
anova(mod1)
```

## Example

The ANOVA corresponds to the sequential tests

1. $y \sim 1$
2. $y \sim \text{block}$
3. $y \sim \text{block} + \text{variety}$

We want to test the blocking effect in this way, within the context of the model as a whole (as opposed to testing it alone against $y$).

## Example

If we examine only the blocking effect, we get a different result

```{r}
anova(lm(yield ~ block))
```

This is because the denonimator in the F-tests is the mean square from the full model, so we lose that here.

(We do not want to do this.)

## Model Terms

Note that the order the terms enter the model doesn't matter because this is an *orthogonal* design. 

```{r}
anova(lm(yield ~ variety+block))
```

## Model Terms

- If the design is not orthogonal, for example if we had one less observation, this would not be the case.
    - the ANOVA would change
    - significance could change
    - etc.
- In this case, we should put the blocking factor into the model *first* so that it's included when we test the treatment factor. 

## Example

Lets remove an observation to see a design that's not orthogonal.

```{r}
anova(lm(yield ~ block+variety,subset=-1, oatvar))
```

## Example

```{r}
anova(lm(yield ~ variety+block,subset=-1, oatvar))
```

Here it only makes a minor difference, but that may not always be the case.

## Example

```{r}
par(mfrow=c(1,2))
plot(mod1, which=c(1,2))
```

## Example

```{r}
TukeyHSD(aov(yield ~ block+variety))$variety
```

## Example: non-additivity test

We might want to confirm interactions aren't important

```{r}
varcoefs <- c(0,coef(mod1)[6:12])
blockcoefs <-c(0,coef(mod1)[2:5])
ab <- rep(varcoefs,each=5)*rep(blockcoefs,8)
h <- update(mod1,.~.+ab)
anova(h)
```

## RCBD versus CRD

- We can measure precision by examining $\hat{\sigma}^2$.
- So we can compare predicion for designs of the same sample size by comparing their $\hat{\sigma}^2$
- To examine *relative efficiency*, we find the ratio of the variances of models with and without the blocking effect. $$\frac{\hat{\sigma}^2_{CRD}}{\hat{\sigma}^2_{RCBD}}$$

## RCBD versus CRD

```{r}
gcrd <- lm(yield ~ variety) # model without block
(summary(gcrd)$sig / summary(mod1)$sig)^2
```

- A CRD would require 66\% more observations to obtain the same level of precision as the RCBD. 
- Increased efficiency not guaranteed for RCBD 
    - use blocks only if there's a strong scientific reason to.

# Latin Squares

## Latin Squares

- Latin Squares are useful when there are two blocking variables.
- Example: risk and gender

## Example Latin Square

- Suppose there are 4 treatments (A-D) and two blocks with 4 levels each.

| | 1 | 2 | 3 | 4 |
| --- | --- | --- | --- | --- |
| I   | A | B | C | D |
| II  | B | D | A | C |
| III | C | A | D | B |
| IV  | D | C | B | A |

- Each treatment assigned to each block exactly once.
    - Each row and column contains all 4 treatments.
    
## Latin Squares 

- We should choose randomly from all possible Latin Square layouts. 
- The R package `magic` can be sued to generate such squares.

## Latin Squares

- Let $\tau$ be the treatment and $\beta$ and $\gamma$ the blocking factors.

The model is $$y_{ij}=\mu + \tau_i + \beta_j + \gamma_{k} + \epsilon_{ijk}$$

- Not all combinations of $i,j,k$ will appear.

## Latin Squares

- Test for treatment effect using F-test on the full model vs a model with the treatment removed.
- The Tukey pairwise CIs are $$\hat{\tau}_l - \hat{\tau}_m \pm \frac{q_{t,(t-1)(t-2)}}{\sqrt{2}}\hat{\sigma}\sqrt{2/t}$$

## Latin Squares vs. RCBD

- Latin Squares more efficient when blocking effects are sizable. 
- Variations:
    - Can be replicated if more runs available.
    - If both block sizes *not* equal to number of treatments, can create Latin Rectangles by adjoining Latin squares.
    - Graeco-Latin squares occasionally used for three blocking variables (but rarely seen in practice).
    
## Latin Squares without Blocks

- Can also be used to compare three treatment factors.
    - Requires $t^2$ runs instead of $t^3$ runs. 
    - Downside: cannot estimate interactions. 
    - This is one type of fractional factorial.
    
## Example

```{r}
data(abrasion)
abrasion
```

## Example

```{r}
xtabs(wear ~ run + position, abrasion)
```

## Example

Latin Square Structure:

```{r}
matrix(abrasion$material, 4, 4)
```

## Example

```{r}
par(mfrow=c(1,3))
with(abrasion, boxplot(wear ~ material, xlab="Material"))
with(abrasion, boxplot(wear ~ run, xlab="Run"))
with(abrasion, boxplot(wear ~ position, xlab="Position"))
```

## Example

```{r}
g <- lm(wear ~ material+run+position, abrasion)
drop1(g, test="F") # test each variable against full model
```

## Example

```{r}
par(mfrow=c(1,2))
plot(g, which=c(1,2))
```

***

```{r results=FALSE}
summary(g)
```

```
            Estimate Std. Error t value Pr(>|t|)    
(Intercept)  254.750      6.187  41.174 1.37e-08 ***
materialB    -45.750      5.534  -8.267 0.000169 ***
materialC    -24.000      5.534  -4.337 0.004892 ** 
materialD    -35.250      5.534  -6.370 0.000703 ***
run2          -2.250      5.534  -0.407 0.698423    
run3          12.500      5.534   2.259 0.064657 .  
run4          -9.250      5.534  -1.671 0.145658    
position2     26.250      5.534   4.743 0.003180 ** 
position3      8.500      5.534   1.536 0.175454    
position4      8.250      5.534   1.491 0.186608    
---

Multiple R-squared:  0.9506,	Adjusted R-squared:  0.8766 
F-statistic: 12.84 on 9 and 6 DF,  p-value: 0.002828
```

## Example

```{r}
qtukey(0.95,4,6)*5.53/sqrt(2)
```

We can them compare pairwise differences to this width:

```{r}
scoefs <- c(0,coef(g)[2:4])
round(outer(scoefs,scoefs,"-"), 2)
```

## Relatively Efficiency

We compare the Latin square to the completely randomized design:

```{r}
lmodr <- lm(wear ~ material,abrasion)
(summary(lmodr)$sig/summary(g)$sig)^2
```

So the Latin square is 3.84 times more efficient than the CRD. 

# Balanced Incomplete Block Designs

## Balanced Incomplete Block Designs

In a complete block design, block size = number treatments. 

What happens when block size < number treatments? 

- Treatments and blocks not orthogonal.
- Some treatment contrasts will not be identifiable from certain block contrasts.
    - *Confounding*: these treatment contrasts cannot be examined. 

## Balanced Incomplete Block (BIB) Designs

- In a BIB design, all pairwise differences are identifiable and have same standard error. 
    - Pairwise differences are more likely to be interesting than other contrasts, so the design is constructed to facilitate this.
    
## Example

Suppose, we have four treatments ($t=4$) A, B, C, D and the block size, $k=3$ and there are $b=4$ blocks.

- Each treatment will appear $r=3$ times.
- Each pair of treatments appears together in a block $\lambda =2$ times. 

## Example

One possibility:

| Block 1 | Block 2 | Block 3 | Block 4 | 
| ------- | ------- | ------- | ------- | 
| A | A | A | B |
| B | B | C | C |
| C | D | D | D |

***

- Treatments ($t$), block size, $k$, and $b$ blocks.
- Each treatment will appear $r$ times.
- Each pair of treatments appears in the *same* block $\lambda$ times. 

For a BIB design, we require

- $b \ge t > k$ 
- $rt = bk = n$
- $\lambda(t-1) = r(k-1)$
    - Number of pairs in a block is $k(k-1)/2$ so total pairs is $bk(k-1)/2$.
    - Number of treatment pairs is $t(t-1)/2$
        - Ratio of these must be $\lambda$

## BIB Designs

Since $\lambda$ must be an integer, a BIB design isn't always possible, even if the first two conditions hold. 

- Ex: $r=4, t=3, b=6, k=2$ so $\lambda=2$ 
    - BIB possible
- Ex: $r=4, t=4, b=8, k=2$ so $\lambda=4/3$ 
    - BIB not possible

## BIB Model

Same as for RCBD: 

$$y_{ij}=\mu + \tau_i + \beta_j + \epsilon_{ij}$$

## Example

- A nutritionist studied the effects of six diets, “a” through “f,” on weight
gain of domestic rabbits.
- From past experience with sizes of litters, it was felt that only three uniform rabbits could be selected from each available litter. 
- There were 10 litters available forming blocks of size 3.

## Example

```{r}
data(rabbit)
head(rabbit)
```

## Example

```{r}
xtabs (gain ~ treat+block, rabbit)
```

- 0 values correspond to no observation. 
- Note BIB structure w/ each pair of diets appearing in same block exactly twice

## Example

```{r message=F}
attach(rabbit); par(mfrow=c(1,2))
boxplot(gain ~ block); boxplot(gain ~ treat)
```


## Example

```{r}
mod1 <- lm(gain ~ block + treat, rabbit) # enter block first!
drop1(mod1, test="F")
```

## Example

- This design is not orthogonal, so we use `drop1` instead of `anova`
- This way, we can examine each term in the presence of the other.

## Example

```{r}
par(mfrow=c(1,2))
plot(mod1, which=c(1,2))
```

## Example: Pairwise CIs

- Since this is not a complete layout, we need to construct the Tukey CIs.

```{r}
coef(summary(mod1))[11:15,]
```

## Example: Pairwise CIs

Compute Tukey CV: 

```{r}
qtukey(0.95, 6, 15) # 6 means to compare, 15 model df
```

## Example: Pairwise CIs

- Std errors are all 2.242

Compute interval width:

```{r}
qtukey(0.95, 6, 15)*2.242/sqrt(2)
```

## Example

```{r}
tcoefs <- c(0, coef(mod1)[11:15])
abs(outer(tcoefs, tcoefs,"-")) > 7.2842 # only e-f different
```
## Example

Relative efficiency of this BIB vs CRD

```{r}
mod.crd <- lm(gain ~ treat, rabbit)
(summary(mod.crd)$sig / summary(mod1)$sig)^2 # way more efficient!
```