---
title: "Example 9.4"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Polynomials

### one predictor variable

```{r}
x <- runif(100, -10, 10)
y <- 3*x + 0.5*x^2 + rnorm(100, 0, 10)
plot(x,y)

mod <- lm(y ~ poly(x,5)); summary(mod)

# lm(y ~ x + I(x^2) + I(x^3))
```

The `poly` function automatically creates orthogonal polynomials, which have nice properties related to numerical stability.

Using `poly`, we can examine the significance for each higher order term and determine where to stop. (Note that if we use another approach to `poly`, we need to refit the model each time we add or remove a higher order term before determining if it is useful.) Based on this output, a quadratic model is best. 

```{r}
mod2 <- lm(y ~ poly(x,2)); summary(mod2)
```

```{r}
plot(x, y)
abline(lm(y ~ x), lwd=2)
points(seq(-12,12,0.1), predict(mod2, data.frame(x=seq(-12,12,0.1))), type="l", col="purple", lwd=2)
```

In general, if a higher order term is significant, we keep all of the lower order terms (even if they're not all statistically significant). 

### step functions

Consider some data generated in the following manner

```{r}
set.seed(0)
n <- 40
x <- seq(-1, 2, length.out = n)
step <- 5*(x < 0.5)
y <- step + rnorm(n)
plot(x, y)
```

How should we fit these data?

```{r}
plot(x, y)
abline(lm(y ~ x))
```

Maybe not our best bet...

Instead, we can fit a step function.

```{r}
mod.step <- lm(y ~ I(x<0.5))
plot(x, y)
lines(x, predict(mod.step), col="red", lwd=2)
```

```{r}
summary(lm(y ~ x))

summary(mod.step)
```

