---
title: "Two Factors with No Replication"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(faraway)
```

## Two-Way ANOVA

We can extend the one-way ANOVA to other settings, such as the two-way ANOVA.

- This is a linear model with two categorical predictors, plus their interaction. 
- ANOVA adds predictors into the model using *sequential hypothesis testing*.

## Two-Way ANOVA

The most general model is:

$$y_{ij} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ij}$$

for two factors, $\alpha$ with $I$ levels and $\beta$ with $J$ levels.

- Note: not all parameters are identifiable. 

## Sequential Hypothesis Testing

| | Null Model | Alternative |
| - | ---- | ----------- |
| 1. | $y \sim 1$ | $y \sim \alpha$ |
| 2. | $y \sim \alpha$ | $y \sim \alpha + \beta$ |
| 3. |  $y \sim \alpha + \beta$ |  $y \sim \alpha + \beta + \alpha\beta$ |
    
## Interaction Effects

- Interaction $(\alpha\beta)_{ij}$ is the part of the mean response not attributable to the additive effects of the factors. 
    - Ex: you might not want to eat straight sugar or drink heavy cream, but whipped cream (a combination of the two) is much more pleasant. 
- Interactions make models harder to interpret, since we can't examine $\alpha$ and $\beta$ independently of each other. 

## Interaction Effects

Consider the following 2x2 examples

|   | Male | Female | | Male | Female |
|---|------|--------|-|------|--------|
|Drug 1| 3 | 5 |Drug 1| 2 | 1 |
|Drug 2| 1 | 2 |Drug 2| 1 | 2 |

The response is some measure of drug performance.

- Left has no interaction, right does.

## Interaction Effects

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(x=c(0,1), y=c(1,2), type='l', ylim=c(1,5), xlab="Sex", ylab="Response", xaxt="n")
points(x=c(0,1), y=c(3,5), type='l')
plot(x=c(0,1), y=c(2,1), type='l', ylim=c(1,2), xlab="Sex", ylab="Response", xaxt="n")
points(x=c(0,1), y=c(1,2), type='l')
```

## Interaction Effects

- When there's an interaction, not a clear way to define the individual effects.
    - Is the gender effect the effect for F? For M? For their average?
- We can also create a model $y_{ijk}=\mu_{ijk} + \epsilon_{ijk}$, treating it as a one-way ANOVA with $I\times J$ total levels. 

## Two-Way ANOVA with One Observation Per Cell

```{r}
data("composite")
composite
```

## One Observation Per Cell

- One observation for each combination of `laser` and `tape`.

We can fit a main effects model:
```{r}
mod1 <- lm(strength ~ laser + tape, data=composite)
```
but this loses sight of possible interactions.

However, if we try to fit an interaction model, parameters = observations

## One Observation Per Cell

```{r}
par(mfrow=c(1,2))
with (composite, interaction.plot (laser, tape, strength, legend=F))
with (composite, interaction.plot (tape, laser, strength, legend=F))
```

## Tukey's Nonadditivity Test

Those don't look like significant interactions, but here's another way to test. We fit the model

$$y_{ij} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ij}$$

and then test if the interaction term is zero.

- Note that this does assume the interaction is multiplicative, which may not always be the case.

## Tukey's Nonadditivity Test

Since these parameters aren't all identifiable, we have to work a little harder. 

```{r}
coef (mod1)
lasercoefs <- rep(c(0,6.5733, 12.2133),3)
tapecoefs <- rep(c(0,4.0367,-1.4800), each=3)
h <- update(mod1, . ~ . + I(lasercoefs*tapecoefs))
```

## Tukey's Nonadditivity Test

```{r}
anova(h)
```

Looks like the interaction is not significant

## One Observation Per Cell

```{r}
anova(mod1)
```

We can then look at the plots and coefficients to see how strength changes with laser power and tape speed.

## One Observation Per Cell

One thing we're not taking into account is that these factors are *ordered*.

```{r results='hide'}
composite$laser <- as.ordered(composite$laser)
composite$tape <- as.ordered(composite$tape)
g <- lm(strength ~ laser + tape, composite)
summary(g) #note the anova itself will look the same
```

***

`            Estimate Std. Error t value Pr(>|t|)    `

`(Intercept)  31.0322     0.5401  57.452  5.5e-07 ***`

`laser.L       8.6361     0.9355   9.231 0.000765 ***`

`laser.Q      -0.3810     0.9355  -0.407 0.704655    `

`tape.L       -1.0465     0.9355  -1.119 0.325944    `

`tape.Q       -3.9001     0.9355  -4.169 0.014045 *  `

`Multiple R-squared:  0.963,	Adjusted R-squared:  0.9259 `

`F-statistic:    26 on 4 and 4 DF,  p-value: 0.004013`
