---
title: 'Logistic Regression'
author: "Dr. Lauren Perry"
output:
  ioslides_presentation: default
  beamer_presentation: default
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR2)
```

# Classification

## Examples of Classification Problems

1. A person arrives at the emergency room with a set of symptoms that could possibly be attributed to one of three medical conditions. Which of the three conditions does the individual have?

2. An online banking service must be able to determine whether or not a transaction being performed on the site is fraudulent, on the basis of the user’s IP address, past transaction history, and so forth.

3. On the basis of DNA sequence data for a number of patients with and without a given disease, a biologist would like to figure out which DNA mutations are deleterious (disease-causing) and which are not.

## Classification Problems

- The data is going to look very similar to the data used in regression. 
- However, now $Y$ will be *categorical*.

## Classification: The `Default` Data

This is a simulated dataset with variables 

- `default`, whether an individual defaulted
- `balance`, credit card balance
- `income`

## Classification: The `Default` Data

```{r}
par(mfrow=c(1,2))
boxplot(Default$balance ~ Default$default, xlab="Default", ylab="Balance")
boxplot(Default$income ~ Default$default, xlab="Default", ylab="Income")
```

## Why not Linear Regression?

Why not just code the categorical output as numbers?

That is, for three diagnoses: `stroke`, `overdose`, and `epileptic seizure`, let 

\[
Y = 
\begin{cases}
1 \quad\text{if stroke} \\
2 \quad\text{if overdose} \\
3 \quad\text{if epileptic seizure}
\end{cases}
\]

What problems might this cause? 

## Why not Linear Regression?

What are we assuming about the diagnoses that's different if we let 

\[
Y = 
\begin{cases}
1 \quad\text{if overdose} \\
2 \quad\text{if stroke} \\
3 \quad\text{if epileptic seizure}
\end{cases}
\]

## What if $Y$ only has two categories? 

Can we use a dummy variable?

Suppose we are only interested in whether a patient has had a stroke. $$Y = I(\text{stroke})$$

## What if $Y$ only has two categories?

- This seems reasonable: we could just run the regression and predict
    - stroke if $\hat{Y} > 0.5$
    - no stroke otherwise
- This works out ok, but we'd like to be able interpret the output as the *probability* of a stroke.
    - However, in this scenario $\hat{Y}$ can take values outside of $[0,1]$.

# Logistic Regression

## Logistic Regression

Logistic regression will allow us to examine the case where the outcome variable has two distinct categories.

-   The outcome $Y$ will be constructed as a dummy variable.
-   Rather than modeling the response directly, logistic regression models the *probability* of falling into a particular category.
    - E.g., $P(\text{default}=\text{yes}|\text{balance})$
-   Then we might use, say, $\hat{Y} = 0.5$ as the cutoff point for predicting one category or the other.

## The Logistic Model

How do we model the relationship between this *probability* and the outcome $X$? $$P(Y=1|X) = p(X)$$

## Example: Default Data

What if we modeled $p(X) = X\beta$

```{r}
summary(Default$default)
def <- as.numeric(Default$default)-1 # 0 is no, 1 is yes
```

## Example: Default Data

```{r}
lm(def ~ balance, Default)
```

- Problem: balances close to 0 or very large result in predicted values outside of [0,1]
- For linear regression, $\hat{y}$ is continuous on $(-\infty, \infty)$.

## Link Functions

How do we get from a probability to an arbitrary real number?

- Suppose we have probabilities $p_i \in [0,1]$
- Link function $g$ takes values $p_i$ and converts them to real numbers $\eta_i = X\beta$
- Now $\eta_i \in (-\infty, \infty)$ and we can run a regression using $\eta$ as the output.

## Link Functions

Properties:

- For these to be useful, $\eta_i=g(p_i)$ needs to have an inverse.
- We require $g$ to be monotone
- For this setting, we need $g^{-1}(\eta)$ to produce values in $[0,1]$.

## Logit Link

In this setting, the most popular choice of link function is the *logit*:

$$\eta = \log\left(\frac{p}{1-p}\right)$$
or, equivalently, 
$$p = \frac{e^{\eta}}{1+e^{\eta}}$$

## The Logit Function

```{r}
eta = seq(-6,6,length.out=100)
p = exp(eta)/(1+exp(eta))
plot(eta, p, type='l')
```

## The Logistic Model

In logistic regression, the logit link takes on the roll of the *logistic function* $$p(X) = \frac{\exp(\beta_0 + \beta_1X)}{1+\exp(\beta_0 + \beta_1X)}$$

## Odds

-   We can rewrite $$\frac{p(X)}{1-p(X)} = \exp(\beta_0 + \beta_1X)$$
    -   The quantity $p(X)/[1-p(X)]$ is called the *odds*
-   Odds can take on any values on $[0, \inf)$.

## Odds

-   Ex: With an odds of 1/4, on avg 1 in 5 people will default.
    -   $p(X) = 0.2 \rightarrow 0.2/(1-0.2) = 1/4$
-   Ex: With an odds of 9, on average 9 in 10 people will default.
    -   $p(X) = 0.9 \rightarrow 0.9/(1-0.9) = 9$

## The Logistic Model

$$p(X) = \frac{\exp(X\beta)}{1+\exp(X\beta)}$$

can be written as

$$\frac{p(X)}{1-p(X)} = \exp(X\beta)$$

and finally as

$$\log\left(\frac{p(X)}{1-p(X)}\right) = X\beta$$

where the LHS is the *log odds* or *logit* function.

## The Logistic Model

$$\log\left(\frac{p(X)}{1-p(X)}\right) = X\beta$$

-   When we run a logistic regression, predicted values are *log odds*.
-   This makes interpretation a little more involved!
-   If we increase $X$ by one unit, the log odds change by $\beta_1$.
    -   Or we can say it multiplies the odds by $e^{\beta_1}$.
    
## Maximum Likelihood

- For linear regression, we used *least squares* to estimate the regression parameters.
- In logistic regression, we use *maximum likelihood*.
    - Select the parameter values with maximize the *likelihood*.
    
## Maximum Likelihood

$$\eta_i = X\beta$$

$$p_i = \frac{e^{\eta_i}}{1+e^{\eta_i}}$$

Show that the log likelihood function is 
$$l(\beta) = \sum_{i=1}^n[y_i\eta_i - \log(1+e^{\eta_i})]$$

## Example: Logistic Regression in R

```{r echo=T}
glm(default ~ balance + income, family=binomial, data=Default)
```

------------------------------------------------------------------------

```{r}
mod1 <- glm(default ~ balance + income, family=binomial, data=Default)
summary(mod1)
```

## Making Predictions

$$\hat{p}(X) = \frac{\exp(X\hat{\beta})}{1+\exp(X\hat{\beta})}$$

and we plug in values for $X$.

-   We can also plug in values for $X$ into the odds or log odds formulations to predict those values.

## Example

```{r results=FALSE}
summary(glm(default ~ ., family=binomial, data=Default))
```

```
Coefficients: Estimate Std. Error z value Pr(>|z|)
(Intercept) -1.087e+01  4.923e-01 -22.080  < 2e-16 ***
studentYes  -6.468e-01  2.363e-01  -2.738  0.00619 **
balance      5.737e-03  2.319e-04  24.738  < 2e-16 ***
income       3.033e-06  8.203e-06   0.370  0.71152
```

-   Negative coefficient suggests students less likely to default than non students.
-   However, a model with *only* `student` has a positive coefficient. Why?

***

![](images/4_3.pdf)

-   Overall average default rates are higher for students.
-   As a function of credit card balance, default rates are *lower* for students.
    - So students are riskier than non-students *on average*, but a student is less risky than a non-student with the same credit card balance. 
    
## Deviance

The deviance allows us to compare models.

- Smaller deviance = better model

$$D = -2\sum_{i-1}^n\hat{p}_i\text{logit}(\hat{p}_i) + \log(1-\hat{p}_i)$$

This is the same idea as $RSS = \sum_{i=1}^n(y_i - \hat{y}_i)^2$ in linear regression.

## Residuals

In logistic regression, $$\sum r_i^2 = D$$

## Assumptions

- Dependent variable is binary
- Observations are independent
- Minimal multicollinearity 
- Linearity of independent variables and log odds
- Typically requires relatively large sample size

## Multinomial Logistic Regression 

What if the response variable has more than two outcomes, $K > 2$?

Select one class to serve as the *baseline*. WLOG, we choose class $K$. Then the model becomes
$$ \text{P}(Y=k | X = x) = \frac{\exp(X\beta_k)}{1 + \sum_{l=1}^{K-1}\exp(X\beta_k)} $$
for $p$ predictor variables and $k = 1, \dots, K-1$ and 
$$ \text{P}(Y=K | X = x) = \frac{1}{1 + \sum_{l=1}^{K-1}\exp(X\beta_l)}. $$

## Multinomial Logistic Regression 

Then for $k = 1, \dots, K-1$, $$\log\left(\frac{\text{P}(Y=k | X = x)}{\text{P}(Y=K | X=x)}\right) = \sum_{i=0}^p \beta_{ki}x_i$$

- Baseline is unimportant from a modeling perspective, but it does serve as the point of comparison for other classes.
- Say `seizure` is our baseline for some medical classification problem and let $k$ represent `stroke`.
    - Then the model gives the log odds of `stroke` versus `seizure`. 
    
## Softmax Coding

Another way to do the estimation for a multinomal logistic regression is, for $k = 1, 2, \dots, K$
$$ \text{P}(Y=k | X = x) = \frac{\exp(\sum_{i=0}^p \beta_{ki}x_i)}{\sum_{l=1}^{K}\exp(\sum_{i=0}^p \beta_{ki}x_i)} $$
which requires the estimation of coefficients for all $K$ classes, instead of $K-1$ classes. 

Then the log odds ratio between *any two classes* $k$ and $k'$ is 
$$ \log\left(\frac{\text{P}(Y=k | X = x)}{\text{P}(Y=k' | X=x)}\right) = \sum_{i=0}^p(\beta_{ki}- \beta_{k'i})x_i$$

## Multinomial Logistic Regression in R

```{r echo=T, eval=F}
library(nnet)
data(penguins, package = "palmerpenguins")
penguins$species <- relevel(penguins$species, ref="Adelie")
mod2 <- multinom(species ~ bill_length_mm + flipper_length_mm 
                 + body_mass_g + sex, data = penguins)
summary(mod2)
```

***

```{r echo=F, eval=F}
library(nnet)
data(penguins, package = "palmerpenguins")
penguins$species <- relevel(penguins$species, ref="Adelie")
mod2 <- multinom(species ~ bill_length_mm + flipper_length_mm + body_mass_g + sex, data = penguins)
summary(mod2)
```

```
Call:
multinom(formula = species ~ bill_length_mm + flipper_length_mm + 
    body_mass_g + sex, data = penguins)

Coefficients:
          (Intercept) bill_length_mm flipper_length_mm body_mass_g   sexmale
Chinstrap   -114.8867      3.6407536        0.02945164 -0.01248832 -10.84175
Gentoo      -141.8286      0.7009754        0.33514983  0.01173179 -15.39547

Std. Errors:
          (Intercept) bill_length_mm flipper_length_mm body_mass_g    sexmale
Chinstrap 0.028765022      1.0261304         0.1411880 0.006587535 0.05226856
Gentoo    0.009824929      0.7616873         0.1561736 0.008052310 0.04375471

Residual Deviance: 7.66202 
AIC: 27.66202
```
