---
title: "Missing Data"
author: "Prof. Lauren Perry"
output:
  ioslides_presentation: default
  beamer_presentation: default
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()

library(faraway)
```

## Missing Data

Missing data occur when some of the values are missing.

-   This is very common!
-   We can remove those observations... but often that means losing a lot of data.

## Types of Missing Data

**Missing Cases**

- Complete case $(x_i, y_i)$ is missing.
- Always missing unobserved data from the population.
- Generally not an issue.
- Can be an issue (bias) if there's some *systematic* reason for our inability to sample certain cases.
    - May be able to make corrections with enough info about mechanism for missingness.
    
## Types of Missing Data

**Incomplete Values**

- These can occur when an experiment ends before all end points are observed.
    - E.g., survival or reliability analysis 
- These can also occur due to individuals dropping out partway through an experiment. 
- It is also possible to have observations where early values are unobserved.
- Such cases are said to be *censored*.

## Types of Missing Data

**Missing Values**

- For a given case, some components are measured by not others. 
- This chapter addresses the missing values problem. 

## Types of Missingness 

**Missing completely at random** (MCAR)

- the probability a value is missing is the same for all cases
    - deleting cases with missing values will remove information, but cause no bias
      
## Types of Missingness 

**Missing at random** (MAR)

- the probability a value is missing depends on some known mechanism
    - need to make some adjustments before deleting

## Types of Missingness 
**Missing not at random** (MNAR)

- the probability a value is missing depends on some unobserved variable or on what value would have been observed
    - eg, people refusing to answer a survey because they have something to hide

Dealing with MNAR is difficult/impossible to do accurately, so we focus on MAR problems. 

## "Fixing" Missing Data

-   Remove observations with (any amount of) missing data.
-   Fill in or *impute* the missing values.
-   Missing value correlation method.
-   Maximum likelihood methods.

## Deletion: Chicago Insurance Data

Some points have been removed at random:

```{r}
data(chmiss)
head(chmiss, 10)
```

***

```{r}
summary(chmiss)
```

***

```{r}
rowSums(is.na(chmiss)) # missing obs per row
```

***

```{r}
image(is.na(chmiss),axes=FALSE,col=gray(1:0))
axis(2, at=0:5/5, labels=colnames(chmiss))
axis(1, at=0:46/46, labels=row.names(chmiss),las=2)
```

***

```{r message=FALSE}
library(DataExplorer)
plot_missing(chmiss, missing_only = TRUE)
```

## Deletion: Chicago Insurance Data

- There is at most one missing obs per row.
- Removing these obs would mean deleting 20 of our 47 observations.

## The Model, Nothing Missing

```{r}
data(chredlin, package="faraway")
modfull <- lm(involact ~ ., chredlin)
summary(modfull)$coefficients
```

## The Model, Data Missing

```{r results=F}
mod1 <- lm(involact ~ ., chmiss)
summary(mod1)
```

```
Coefficients:
Estimate Std. Error t value Pr(>|t|)    
(Intercept) -1.116483   0.605761  -1.843 0.079475 .  
race         0.010487   0.003128   3.352 0.003018 ** 
fire         0.043876   0.010319   4.252 0.000356 ***
theft       -0.017220   0.005900  -2.918 0.008215 ** 
age          0.009377   0.003494   2.684 0.013904 *  
income       0.068701   0.042156   1.630 0.118077    

Residual standard error: 0.3382 on 21 degrees of freedom
  (20 observations deleted due to missingness)
Multiple R-squared:  0.7911,	Adjusted R-squared:  0.7414 
F-statistic: 15.91 on 5 and 21 DF,  p-value: 1.594e-06
```

## Deletion: Chicago Insurance Data

- `lm` function in R automatically removes any observations with anything missing. 
- Reduction in data has resulted in higher standard errors.
    - (Estimates are less precise.)
- Deleting observations is easy and avoids possible bias introduced by other methods.
    - Downside is information loss. 

# Single Imputation

## Mean Imputation

```{r}
cmeans <- apply(chmiss, 2, mean, na.rm=TRUE)
cmeans
mchm <- chmiss
for(i in c(1, 2, 3, 4, 6)){ # 5 is the outcome variable
  mchm[is.na(chmiss[,i]),i] <- cmeans[i]
}
```

Note: we don't impute values for the outcome because this is the variable we're trying to model.

## Refitting the Model

```{r results=F}
mod2 <- lm(involact ~ ., mchm)
summary(mod2)
```

```
Coefficients:
             Estimate Std. Error t value Pr(>|t|)  
(Intercept)  0.070802   0.509453   0.139  0.89020   
race         0.007117   0.002706   2.631  0.01224 * 
fire         0.028742   0.009385   3.062  0.00402 **
theft       -0.003059   0.002746  -1.114  0.27224   
age          0.006080   0.003208   1.895  0.06570 . 
income      -0.027092   0.031678  -0.855  0.39779  

Residual standard error: 0.3841 on 38 degrees of freedom
  (3 observations deleted due to missingness)
Multiple R-squared:  0.682,	Adjusted R-squared:  0.6401 
F-statistic:  16.3 on 5 and 38 DF,  p-value: 1.409e-08
```

## Refitting the Model

- Notice that theft and age are no longer significant.
- Regression coefs all closer to zero than before.

## Mean Imputation

- Bias/Variance Tradeoff
    - Reduction in variance
    - Increase in bias (coefficients toward zero)
    - May not be worth it unless the proportion of filled values is small
    
## Using Regression Methods

We can also use regression-based methods to *predict* the missing values of the covariates:

```{r}
chmiss[is.na (chmiss$fire) , ]
race.miss <- lm (fire ~ race + theft + age + income, chmiss)
predict(race.miss, chmiss[is.na(chmiss$fire),])
```

## Using Regression Methods

How do our predicted values compare to the actual values? 

```{r}
data(chredlin)
chredlin$fire[is.na(chmiss$fire)]
```

- This method also introduces some bias toward zero in coefs.
- Tends to reduce variance.
- This method works better the more collinear the predictors are.

# Multiple Imputation

## Multiple Imputation

Goal: maintain information while reducing the bias caused by single imputation methods

- Single imputation tends to reduce variability, increasing bias. 
- Idea: re-include that error variation over multiple iterations.

## Multiple Imputation

- One approach is implemented in the `Amelia` package in R.
    - A bootstrap EM (expectation-maximization) algorithm
- Generally assumes data is multivariate normal
    - May need to log-transform heavily skewed variables.
    
## Multiple Imputation

```{r results=FALSE, message=FALSE}
require(Amelia)
set.seed(0)
chimp <- amelia(chmiss, m=25)
```

- `m` is the number of imputations. 
- We should include the entire dataframe, even if we do not intend to use all the variables in our model.
- Next, we will fit a linear model to each of the 25 imputed datasets and store the results.

## Multiple Imputation

```{r}
betas <- NULL
ses <- NULL
for(i in 1:chimp$m){
  lmod <- lm(involact ~ race+fire+theft+age, chimp$imputations[[i]])
  betas <- rbind(betas, coef(lmod))
  ses <- rbind(ses, coef(summary(lmod))[,2])
}
```

## Multiple Imputation

We can combine the results to get estimates of the coefficients and their standard errors.

- Average over the imputations for $\beta_j$: $$\hat{\beta}_j = \frac{1}{m}\sum_{i}\hat{\beta}_{ij}$$
- Combined standard errors calculated as $$s_j^2 = \frac{1}{m}\sum_i s_{ij}^2 + \text{var}(\hat{\beta}_j)(1+1/m)$$

## Multiple Imputation

The `mi.meld` function does this for us:

```{r}
(cr <- mi.meld(q=betas,se=ses))
```

## Multiple Imputation

And the t-statistics are

```{r}
cr$q.mi/cr$se.mi
```