---
title: "Two Factors with an Interaction"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(faraway)
```

## Example: Warp Breaks

```{r}
data("warpbreaks")
par(mfrow=c(1,2))
plot(breaks ~ wool, warpbreaks); plot(breaks ~ tension, warpbreaks)
```

## Example: Warp Breaks

```{r}
par(mfrow=c(1,2))
with(warpbreaks,interaction.plot(wool,tension,breaks))
with(warpbreaks,interaction.plot(tension,wool,breaks))
```

## Initial Diagnostics

```{r}
lmod <- lm(breaks ~ wool*tension, warpbreaks)
plot(lmod, 1)
```

## Transformation

Nonconstant variance, since response is count variable we will transform using a square root.

```{r}
lmod <- lm(sqrt(breaks) ~ wool*tension, warpbreaks)
plot(lmod,1)
```

## Two-Way ANOVA

```{r}
anova(lmod)
```

## Another Way to Examine Interactions

- May be easier to combine the factor levels into one combined `treatment`
    - Will have a number of levels equal to the product of the two factor levels.

*** 

```{r results=FALSE}
lmod <- lm(sqrt(breaks) ~ -1 + wool:tension, warpbreaks)
summary(lmod)
```

```
               Estimate Std. Error t value Pr(>|t|)    
woolA:tensionL   6.5476     0.3266   20.05   <2e-16 ***
woolB:tensionL   5.2381     0.3266   16.04   <2e-16 ***
woolA:tensionM   4.8259     0.3266   14.78   <2e-16 ***
woolB:tensionM   5.2987     0.3266   16.22   <2e-16 ***
woolA:tensionH   4.8564     0.3266   14.87   <2e-16 ***
woolB:tensionH   4.3022     0.3266   13.17   <2e-16 ***
---

Residual standard error: 0.9799 on 48 degrees of freedom
Multiple R-squared:  0.9697,	Adjusted R-squared:  0.9659 
F-statistic: 255.8 on 6 and 48 DF,  p-value: < 2.2e-16
```

## Tukey's HSD 

```{r results=FALSE}
TukeyHSD(aov(breaks ~ wool:tension, warpbreaks))
```

```
               diff       lwr        upr     p adj
B:L-A:L -16.3333333 -31.63966  -1.027012 0.0302143
A:M-A:L -20.5555556 -35.86188  -5.249234 0.0029580
B:M-A:L -15.7777778 -31.08410  -0.471456 0.0398172
A:H-A:L -20.0000000 -35.30632  -4.693678 0.0040955
B:H-A:L -25.7777778 -41.08410 -10.471456 0.0001136
A:M-B:L  -4.2222222 -19.52854  11.084100 0.9626541
B:M-B:L   0.5555556 -14.75077  15.861877 0.9999978
A:H-B:L  -3.6666667 -18.97299  11.639655 0.9797123
B:H-B:L  -9.4444444 -24.75077   5.861877 0.4560950
B:M-A:M   4.7777778 -10.52854  20.084100 0.9377205
A:H-A:M   0.5555556 -14.75077  15.861877 0.9999978
B:H-A:M  -5.2222222 -20.52854  10.084100 0.9114780
A:H-B:M  -4.2222222 -19.52854  11.084100 0.9626541
B:H-B:M -10.0000000 -25.30632   5.306322 0.3918767
B:H-A:H  -5.7777778 -21.08410   9.528544 0.8705572
```

