---
title: "Example 3.4"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Consider the full model for the savings data.

```{r}
library(faraway)
data(savings)

mod1 <- lm(sr ~ ., data=savings)
sumry1 <- summary(mod1)
```

### Univariate Confidence Intervals

Let's construct a 95\% CI for $\beta_{pop75}$:

```{r}
sumry1

est <- mod1$coefficients[3]
cv <- qt(0.975,45)
se <- sumry1$coefficients[3,2]

est + c(-1,1)*cv*se
```

```{r}
summary(mod1)
```

We can also, as usual, do this more conveniently:

```{r}
confint(mod1)
```

### Joint Confidence Regions

```{r}
# install.packages("ellipse")
library(ellipse)
```

```{r}
# plot the ellipse for pop15 and pop75
plot(ellipse(mod1, c(2,3)), type="l", xlim=c(-1,0.5))
# add origin and point estimates of estimates
points(0,0)
points(coef(mod1)[2], coef(mod1)[3], pch=18)
```

Origin is outside of ellipse, so we can reject $H_0: \beta_{pop15} = \beta_{pop75} = 0$. 

Note that the joint nature of the ellipse means that the univariate CIs are not at all equivalent to what's happening here:

```{r}
plot(ellipse(mod1, c(2,3)), type="l", xlim=c(-1,0.5))
# add origin and point estimates of estimates
points(0,0)
points(coef(mod1)[2], coef(mod1)[3], pch=18)
# add one-way CIs to plot
abline(v=confint(mod1)[2,], lty=2)
abline(h=confint(mod1)[3,], lty=2)
```

```{r}
cor(savings$pop15, savings$pop75)
```

So the joint test is going to be more meaningful!

```{r}
summary(mod1, corr=TRUE)$corr
```

Notice that the correlation between `pop15` and `pop75` is negative, but the correlation between those two predictors is positive. 

Why? They're providing a lot of redundant information, so the more "work" one does, the less the other needs to do. 