---
title: "Two Factors with Replication"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(faraway)
```

## Two-Way ANOVA

Two-way ANOVA with more than one observation per cell:

Consider the setting where $n_{ij}>1$ is the same for all $i,j$.

- We call this design *orthogonal*

Our model $$y_{ij} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ij}$$ is now identifiable and can be examined in the usual ways.

## Two-Way ANOVA

```{r warning=FALSE}
data(pvc); attach(pvc); par(mfrow=c(1,2))
stripchart (psize ~ resin, xlab="Particle size", ylab="Resin railcar")
stripchart (psize ~ operator, xlab="Particle size", ylab="0perator")
```

## Two-Way ANOVA

It can be hard to tell random noise vs. actual interactions.

```{r echo=FALSE}
par(mfrow=c(1,2))
interaction.plot (operator, resin, psize, legend=F)
interaction.plot (resin, operator, psize, legend=F)
```

## Two-Way ANOVA

```{r}
mod2 <- lm(psize ~ operator*resin, pvc)
anova(mod2)
```

## Two-Way ANOVA

```{r}
anova(lm(psize ~ operator+resin, pvc))
```

## ANOVA Diagnostics

```{r}
par(mfrow=c(1,2))
plot(mod2, which=c(2,1))
```

## ANOVA Diagnostics

- Some evidence of outliers.
- Non-constant error variances.
- Symmetry is because, for the full model, the fitted value for each combination of factors is the mean of the two replicates.

## ANOVA Diagnostics

```{r}
mod4 <- lm(psize ~ operator+resin, pvc[-45,]) # remove outlier
anova(mod4)
```

## Pairwise Comparisons

```{r eval=FALSE}
TukeyHSD(aov(mod4))
```

```
$operator
         diff       lwr        upr     p adj
2-1 -0.262500 -1.073718  0.5487178 0.7114176
3-1 -1.777083 -2.601711 -0.9524561 0.0000184
3-2 -1.514583 -2.339211 -0.6899561 0.0001985

$resin
          diff         lwr        upr     p adj
2-1 -1.0333333 -2.77520402  0.7085374 0.5566387
3-1 -5.8000000 -7.54187069 -4.0581293 0.0000000
4-1 -6.1833333 -7.92520402 -4.4414626 0.0000000
```
etc.

## A Few Notes

- The factors in these models are *fixed effects*.
    - If factors (operators and resin) were randomly selected from larger population, they would be *random effects*, which are a little more complicated to deal with.
    
## A Few Notes

- It is important that observations be genuine replications. 
    - If they are correlated, the analysis will need to be adjusted.
    - Repeated measurements are *not* replications and will be correlated. 
        - Repeated measures ANOVA, etc.