---
title: "Complete Regression Example"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Insurance Redlining

Insurance redlining is the practice of refusing to issue insurance to certain types of people or within some geographic area. 

- Insurance companies may this in a way that is reasonable.
    - Ex: refusing to insure someone with multiple DUIs.

## Insurance Redlining    

- In this dataset, however, our goal is to look for examples of discriminatory insurance redlining in late 1970s Chicago. 
- Examine FAIR plan policies by zip code between Dec 1977 and May 1978. 
    - This plan was offered to homeowners who had been rejected by the voluntary market, so it's a good proxy for examining insurance refusals.

## The Variables

- `race`: racial composition in percentage of minority
- `fire`: fires per 100 housing units
- `theft`: thefts per 1000 population
- `age`: percentage of housing units built before 1939
- `involact`: new FAIR plan policies and renewals per 100 housing units
- `income`: median family income in thousands of dollars
- `side`: north or south side of Chicago

## Ecological Correlation

The *ecological fallacy* is concluding that a correlation between two variables holds at the individual level.

- Ex: countries with higher dietary fat intakes tend to have higher rates of breast cancer, but *individuals* with higher fat diets aren't necessarily at a higher risk of breast cancer. 

In the Chicago dataset, we only have aggregated data, *not* individual data, so we will need to be mindful of this in our analysis.

## Initial Data Analysis

```{r}
data(chredlin, package="faraway")
attach(chredlin)
str(chredlin)
```

## What can we observe?

```{r}
summary(chredlin)
```
## What can we observe?

- Race is a percent, so there's a good amount of variability here.
- Skew in `theft` and `income`
- Response variable has a lot of 0s, which isn't a great sign for our model assumptions.

## Initial Data Analysis

```{r}
round(cor(chredlin[,1:6]),3)
```

## Graphical Summaries

```{r eval=FALSE}
par(mfrow=c(2,3))
# histograms for numeric variables
plot(race, involact)
plot(fire, involact)
plot(theft, involact)
plot(age, involact)
plot(income, involact)
# boxplot for categorical
plot(involact~side)
```

## Graphical Summaries

```{r echo=FALSE}
par(mfrow=c(2,3))
plot(race, involact)
plot(fire, involact)
plot(theft, involact)
plot(age, involact)
plot(income, involact)
plot(involact~side)
```

## What can we observe?

- Definitely some fairly strong relationships here. 
- Some potential outliers/influential points.

## Initial Analysis

```{r eval=FALSE}
summary(lm(involact ~ race,chredlin))
```

```
Coefficients:
            Estimate Std. Error t value Pr(>|t|)    
(Intercept) 0.129218   0.096611   1.338    0.188    
race        0.013882   0.002031   6.836 1.78e-08 ***
---

Multiple R-squared:  0.5094,	Adjusted R-squared:  0.4985 
F-statistic: 46.73 on 1 and 45 DF,  p-value: 1.784e-08
```

## Initial Analysis

Zip codes with a higher proportion of minorities are being rejected for home insurance at a higher rate.

Question:

- Is this due to discrimination, or is it due to differential risk factors?

***

```{r}
par(mfrow=c(1,2))
plot(race, fire); abline(lm(fire~race))
plot(race, theft); abline(lm(theft~race))
```

## Initial Analysis

- There is some correlation bw fire rate and prop of minorities.
- Outlier makes it harder to see correlation bw theft and prop of minorities.

Question:

- If we control for these risk factors by including them in the model, is race still significant?

## Full Model

```{r}
lmod <- lm(involact ~ race + fire + theft + age + log(income) + side, chredlin)
```

- Note we are using `log(income)` 
    - This should help deal with some of the skew in this variable.
    - Makes sense to consider income on a multiplicative scale.
        - ($1000 worth a lot more to us than to a billionaire)
- We examine the model summary next, but we should avoid commenting on it too much until we've done our diagnostics.

***

```{r eval=FALSE}
summary(lmod)
```

```
Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept) -1.190806   1.114419  -1.069  0.29168    
race         0.009330   0.002847   3.277  0.00217 ** 
fire         0.039994   0.008936   4.475 6.19e-05 ***
theft       -0.010227   0.002899  -3.528  0.00107 ** 
age          0.008423   0.002857   2.948  0.00532 ** 
log(income)  0.343419   0.405406   0.847  0.40198    
sides        0.016287   0.124922   0.130  0.89692    
---

Residual standard error: 0.3386 on 40 degrees of freedom
Multiple R-squared:  0.7518,	Adjusted R-squared:  0.7146 
F-statistic:  20.2 on 6 and 40 DF,  p-value: 1.059e-10
```

## Diagnostics

```{r}
par(mfrow=c(1,2))
plot(lmod, 1:2)
```

## Diagnostics

- The diagonal line in the residuals is caused by the large number of 0s in the response. 
    - When $y = 0$, the residual $\hat{e} = −\hat{y} = -x^T\hat{\beta}$
- Both of these plots look reasonable so, despite earlier concerns, we can proceed.

## Diagnostics

```{r}
plot(lmod, 5)
```

## Diagnostics

```{r}
chredlin[c("60607","60610"),]
```

## Diagnostics

These are high theft, high fire rate zip codes.

```{r}
match(c("60607", "60610"),row.names(chredlin))
lmode <- lm(involact ~ race + fire + theft + age + log(income),
            chredlin,subset=-c(6,24))
```

***

```{r eval=FALSE}
summary(lmode)
```
```
Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept) -0.576737   1.080046  -0.534   0.5964    
race         0.007053   0.002696   2.616   0.0126 *  
fire         0.049647   0.008570   5.793    1e-06 ***
theft       -0.006434   0.004349  -1.479   0.1471    
age          0.005171   0.002895   1.786   0.0818 .  
log(income)  0.115703   0.401113   0.288   0.7745    
---

Multiple R-squared:  0.8041,	Adjusted R-squared:  0.779 
F-statistic: 32.01 on 5 and 39 DF,  p-value: 8.192e-13
```

## Diagnostics

- Theft and age are no longer significant.
- Our primary variable of interest, `race` remains significant.
    - That conclusion is robust to the removal of one or two extreme points. 

## Transformations

We should now spend some time experimenting with transformations of the predictors. 

```{r}
par(mfrow=c(1,2))
termplot(lmod, partial.resid=TRUE, terms=1:2)
```

## Sensitivity Analysis

How robust is our conclusion?

- If we selected a different subset of covariates, would `race` still be statistically significant?
- We examine all possible subsets of predictors and examine the p-values for each `race` coefficient
    - I have sorted the p-values from largest to smallest, as they will not all fit on the slide.
    - See the RMD file for the code and complete output.

## Sensitivity Analysis

```{r echo=FALSE}
listcombo <- unlist(sapply(0:5,function(x) combn(5, x, simplify= FALSE)),
                    recursive=FALSE)
predterms <- lapply(listcombo, function(x) 
  paste(c("race",c("fire","theft","age","log(income)","side")[x]),collapse="+"))
coefm <- matrix(NA,32,2)
for(i in 1:32){
  lmi <- lm(as.formula(paste("involact ~ ",predterms [[i]])), data=chredlin)
  coefm[i,] <- summary(lmi)$coef[2,c(1 ,4)]
}
rownames(coefm) <- predterms
colnames(coefm) <- c("beta","pvalue")
round(coefm[order(coefm[,2], decreasing=T),],4)
```

## Sensitivity Analysis 

- All of the p-values associated with the `race` variable are significant
    - We can be confident that the effect is not sensitive to our choice of predictors.
    - It *could* be sensititive to some other predictor not in our data. 
    
## Final Model

In this case, the primary modeling goal is examining the significance of `race` in insurance denials.

- We probably want to keep other risk factors in the model, regardless of their significance. 

Regression is challenging in part because there usually isn't a "right" answer! 