---
title: "Example 4"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Confidence Intervals for Predicted Values

```{r}
library(faraway)
data("gala")

mod1 <- lm(Species ~ Area + Elevation + Nearest + Scruz + Adjacent, gala)
```

Suppose we want to predict the number of tortoise species on an island with 

- Area = 0.08
- Elevation = 93
- Nearest = 6.0
- Scruz = 12.0
- Adjacent = 0.34

```{r}
x0 <- c(1, 0.08, 93, 6.0,  12.0, 0.34)
y0 <- sum(x0*coef(mod1))
y0
```

Now we can work on our confidence intervals. Suppose we want to make a prediction for a specific island which was not measured in the original data.

First, we will compute the confidence interval (for a mean response)

```{r}
x <- model.matrix(mod1)
xtxi <- solve(t(x)%*%x)

summary(mod1)
sigma <- 60.98
df <- 24
```

```{r}
bm <- as.vector(sqrt(x0 %*% xtxi %*% x0) * qt(0.975, df) * sigma)
bm

y0 + c(-1,1)*bm
```

Now let's compute the *prediction* interval (for a specific response)

```{r}
bm2 <- as.vector(sqrt(1 + x0 %*% xtxi %*% x0) * qt(0.975, df) * sigma)
bm2

y0 + c(-1,1)*bm2
```

We can also accomplish this using the `predict` function, but it requires the input to be a data frame that matches that of the original $X$ matrix, so takes a little extra work to set up:

```{r}
x.new <- data.frame(Area = 0.08, Elevation = 93, Nearest = 6,
                    Scruz = 12, Adjacent = 0.34)
predict(mod1, x.new)
# or we can get more info using
predict(mod1, x.new, se=TRUE)
```

We can also use this to generate confidence and prediction intervals

```{r}
predict(mod1, x.new, interval="confidence")

predict(mod1, x.new, interval="prediction")
```

We can also generate confidence bands based on varying a specific variable input (and holding everything else constant):

```{r}
grid <- seq(0, 100, 1)
p <- predict (mod1, data.frame(Area=0.08, Elevation=93, Nearest=grid, Scruz=12, Adjacent=0.34),
               se=T, interval="confidence")
matplot(grid, p$fit, lty=c (1,2,2), type='l', xlab="Nearest",ylab="Species")
rug(gala$Nearest)
```

What happens when we predict values outside the range of the original data?

```{r}
summary(gala$Nearest)
```

