---
title: "Example 3.2: Hypothesis Tests to Compare Models"
author: "Lauren Perry"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Full Model

```{r}
library(faraway)
data(savings)
head(savings)
```

Full model:

```{r}
mod1 <- lm(sr ~ ., data=savings)
summary(mod1)
```

The summary output automatically gives us the results for the test of $H_0: \beta_1=\beta_2=\beta_3=\beta_4=0$

We can also perform this test directly.

```{r}
tss <- sum((savings$sr - mean(savings$sr))^2)
tss

rss <- deviance(mod1)
rss

df.residual(mod1)

fstat <- ((tss-rss)/4) / (rss/df.residual(mod1))
fstat

1-pf(fstat, 4, df.residual(mod1))
```

Which we can verify using the output from the `summary` command, above.

## One Predictor

Let's test if $\beta_1=0$.

```{r}
summary(mod1)
```

We can also do this using the F-test approach:

```{r}
mod2 <- lm(sr ~ . -pop15, data=savings) #model with beta1 removed

rss2 <- deviance(mod2)
# q = p-1, so p-q=1
fstat <- (deviance(mod2) - deviance(mod1))/(deviance(mod1)/df.residual(mod1))

1-pf(fstat, 1, df.residual(mod1))
```

The t test statistic is just the square root of the F test statistic:

```{r}
sqrt(fstat)

tstat <- summary(mod1)$coef[2,3] # t-value from the summary table
tstat
2*(1-pt(sqrt(fstat), 45))
```

### Nested Models

A convenient way to compare nested models in R

```{r}
anova(mod2, mod1)
```

### Testing a pair of predictors

It's important to note that we cannot use the t-test outlined above to test multiple predictors *at the same time*! Why?

Instead, we can test multiple predictors jointly using nested models. Both their p-values are high in the full model, so let's test if it makes sense to remove both `pop75` and `dpi`.

```{r}
mod3 <- lm(sr ~ pop15 + ddpi, savings)
anova(mod3, mod1)
```

In this case, the F-test is not significant, so we fail to reject the hypothesis that $\beta_2=\beta_3=0$. We might then move forward with `mod3` as our preferred model. 

## Testing a Subspace

Let's say we want to test whether the effect of old and young people on savings rate is the same: $$H_0: \beta_{pop15}=\beta_{pop75}$$

Then our null model takes the form $$y=\beta_0 +\beta_{dep}(pop15 + pop75)+\beta_{dpi}dpi + \beta_{ddpi}ddpi + \epsilon$$

```{r}
mod4 <- lm(sr ~ I(pop15+pop75) + dpi + ddpi, data=savings) 
anova(mod4, mod1)
```

Note the function `I` forces R to evaluate that sum before it goes into the model.

Since the p-value is 0.21, we fail to reject the null, so it is reasonable to group old and young people together in the context of this model. 

### Testing Fixed Coefficient Values

We can also extend this test to examine whether a specified coefficient can be set to a particular value. $$H_0: \beta_{ddpi}=0.5$$

So the null model is $$y=\beta_0 +\beta_{pop15}pop15 + \beta_{pop75}pop75+\beta_{dpi}dpi + 0.5ddpi + \epsilon$$

A fixed coefficient is called an *offset*

```{r}
mod5 <- lm(sr ~ pop15 + pop75 + dpi + offset(0.5*ddpi), savings)
anova(mod4, mod1)
```

Since our p-value is large, we fail to reject $H_0$. 

We can also test this using a t-statistic $$t = (\hat{\beta}-c)/SE(\hat{\beta})$$ where $c$ is the point hypothesis.

```{r}
summary(mod1)
tstat <- (0.4096949 - 0.5)/0.1961971
2*pt(tstat, 45)
```

