---
title: "Example 6.3"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Added Variable Plots

Using the `savings` data, we will construct an added variable plot for `pop15`.

```{r}
library(faraway)
data("savings")

d <- residuals(lm(sr ~ pop75 + dpi + ddpi,savings))
m <- residuals(lm(pop15 ~ pop75 + dpi + ddpi,savings))
plot(m,d,xlab="pop15 residuals",ylab="Savings residuals")
```

Let's try adding a regression line to this plot.

```{r}
coef(lm(d ~ m))
lmod <- lm(sr ~ pop15+pop75+dpi+ddpi,savings)
coef(lmod)

plot(m,d,xlab="pop15 residuals",ylab="Savings residuals")
abline(0,coef(lmod)['pop15'])
```

## Partial Residual Plots

```{r}
termplot(lmod, partial.resid=TRUE, terms=1)
```

Note: this centers $x_i$ so that $\hat{\epsilon}} + \hat{\beta}}_i(x_i - \bar{x}_i)$ have mean zero.

There's some kind of grouping structure here. Let's investigate that a little bit.

```{r}
mod1 <- lm(sr ~ pop15+pop75+dpi+ddpi,savings,subset=(pop15 > 35))
mod2 <- lm(sr ~ pop15+pop75+dpi+ddpi,savings,subset=(pop15 < 35))
summary(mod1)
summary(mod2)
```

The model is not significant for `pop15` over 35\%, but is for under 35! 

What's happening here? 

```{r}
rownames(savings[savings$pop15 > 35,])

rownames(savings[savings$pop15 < 35,])
```

In general, developing nations tend to have a higher proportion of children than developed nations. (This dataset is from the 1960s, so these may not match up with our expectations in terms of "developing nations".)

### Extra stuff with ggplot2

```{r}
savings$status <- ifelse(savings$pop15 > 35, "young", "old")
require(ggplot2)
ggplot(savings,aes(x=ddpi,y=sr,shape=status))+geom_point()
ggplot(savings,aes(x=ddpi,y=sr))+geom_point()+facet_grid(~ status)+
stat_smooth(method="lm")
```

