---
title: "Larger Factorial Experiments"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(faraway)
```

## Larger Factorial Experiments

Suppose we have three or more factors $\alpha, \beta, \gamma, \dots$ each with some number of levels $l_{\alpha}, l_{\beta}, l_{\gamma}, \dots$.

- A *full factorial* experiment has at least one run for each combination of factor levels (equal to $l_{\alpha}\times l_{\beta}\times l_{\gamma}\times\dots$)
    - This can easily be a very large number of treatments!
    - Could also include all possible interactions, higher order terms, etc.
    - Usually not performed for more than three or four factors.

## Larger Factorial Experiments

- If no interactions significant, we get several one-way experiments for the cost of one.
- Efficient with experimental resources.
- Often done without replication due to cost concerns.
    - May need to assume some higher order interactions are zero in order to have enough degrees of freedom to examine lower level effects.
    
## Fractional Factorials

Fractional factorials use only a fraction of the number of runs done in a full factorial experiment.

## Fractional Factorials

Consider an experiment with 7 factors, each with 2 levels:

| Effect | mean | main | 2-way | 3-way | 4 | 5 | 6 | 7 |
|---|---|---|---|---|---|---|---|---|
| Parameters | 1 | 7 | 21 | 35 | 35 | 21 | 7 | 1 |

- Assuming higher order interactions are negligible will save us from doing $2^7=128$ runs. 
- 8 runs would allow us to estimate main effects.
- Hard to find a good design to examine all 2-way effects without a lot of runs.

## Fractional Factorials

Goal: estimate as many parameters as possible with as few data points as possible.

- May not have many degrees of freedom left over.
- $\sigma^2$ must be small for this to work (to distinguish significant effects).

## Example: Speedometer cables

- 15 factors labeled `a` through `o`
- Two levels per factor, high and low
  - A `+` indicates high and `-` indicates low
- Design with only 15 runs used to estimate overall mean with 15 main effects.
- Assume no interactions

```{r eval=FALSE}
data(speedo)
speedo # two levels of each factor: +/-
```

## Example

```{r echo=FALSE}
data(speedo)
speedo # two levels of each factor: +/-
```

***

```{r results=FALSE}
mod5 <- lm(y ~ ., speedo)
summary(mod5) # n=p
```

```
Residuals:
ALL 16 residuals are 0: no residual degrees of freedom!

Coefficients:
              Estimate Std. Error t value Pr(>|t|)
(Intercept)  0.5825000        NaN     NaN      NaN
h-          -0.0621875        NaN     NaN      NaN
d-          -0.0609375        NaN     NaN      NaN
...
g-           0.1403125        NaN     NaN      NaN
o-          -0.0059375        NaN     NaN      NaN

Residual standard error: NaN on 0 degrees of freedom
Multiple R-squared:      1,	Adjusted R-squared:    NaN 
F-statistic:   NaN on 15 and 0 DF,  p-value: NA
```
***

```{r results=FALSE}
model.matrix(mod5) # + is coded as 0; - as 1 bc of ASCII alphabet
```

```
   (Intercept) h- d- l- b- j- f- n- a- i- e- m- c- k- g- o-
1            1  1  1  0  1  0  0  1  1  0  0  1  0  1  1  0
2            1  0  1  1  1  1  0  0  1  1  0  0  0  0  1  1
3            1  1  0  1  1  0  1  0  1  0  1  0  0  1  0  1
4            1  0  0  0  1  1  1  1  1  1  1  1  0  0  0  0
5            1  1  1  0  0  1  1  0  1  0  0  1  1  0  0  1
6            1  0  1  1  0  0  1  1  1  1  0  0  1  1  0  0
7            1  1  0  1  0  1  0  1  1  0  1  0  1  0  1  0
8            1  0  0  0  0  0  0  0  1  1  1  1  1  1  1  1
9            1  1  1  0  1  0  0  1  0  1  1  0  1  0  0  1
10           1  0  1  1  1  1  0  0  0  0  1  1  1  1  0  0
11           1  1  0  1  1  0  1  0  0  1  0  1  1  0  1  0
12           1  0  0  0  1  1  1  1  0  0  0  0  1  1  1  1
13           1  1  1  0  0  1  1  0  0  1  1  0  0  1  1  0
14           1  0  1  1  0  0  1  1  0  0  1  1  0  0  1  1
15           1  1  0  1  0  1  0  1  0  1  0  1  0  1  0  1
16           1  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0
```

## Example

Can't do usual F-tests, so need different approach.

- Suppose no significant effects and errors normally distributed.
    - Estimated effects would be linear combos of errors and so also normal.
- We can make a normal QQ plot of main effects.
    - Idea: outliers represent something interesting happening (significant effects).
        
## Example

```{r}
par(mfrow=c(1,2))
qqnorm(coef(mod5)[-1], pch=names(coef(mod5)[-1]))
halfnorm(coef(mod5)[-1],labs=names(coef(mod5)[-1]))
```

## Example

So we might further examine effects of `e` and `g`.