---
title: "Model Selection"
author: "Lauren Perry"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(faraway)
```

***

Goal: select the ``best`` subset of predictors for a regression model. 

- Want to select *parsimonious* models
    - Balance between informativeness and complexity.
    
# Testing-Based Procedures
=
## Backward Elimination 

1. Start with all predictors in the model. 
2. Remove the predictor with the highest p-value greater than $\alpha$.
3. Refit the model with the remaining predictors.
4. Repeat steps 2 and 3 until all p-values are less than $\alpha$. 

## Forward Selection

1. Start with no variables in the model.
2. For all predictors not in the model, check their p-values if they are added to the model.
3. Choose the predictor with the smallest p-value less than $\alpha$ and add it to the model. 
4. Repeat steps 2 and 3 until no new predictors can be added.

## Stepwise Regression

- This describes any combination of backward elimination and forward selection.
- At any stage, a variable may be added or removed from the model based on p-values. 

## Example: Backward Elimination 

```{r echo=TRUE, results='hide'}
data(state)
statedata <- data.frame(state.x77,row.names=state.abb)
lmod <- lm(Life.Exp ~ ., statedata)
summary(lmod)
```
```
Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept)  7.094e+01  1.748e+00  40.586  < 2e-16 ***
Population   5.180e-05  2.919e-05   1.775   0.0832 .  
Income      -2.180e-05  2.444e-04  -0.089   0.9293    
Illiteracy   3.382e-02  3.663e-01   0.092   0.9269    
Murder      -3.011e-01  4.662e-02  -6.459 8.68e-08 ***
HS.Grad      4.893e-02  2.332e-02   2.098   0.0420 *  
Frost       -5.735e-03  3.143e-03  -1.825   0.0752 .  
Area        -7.383e-08  1.668e-06  -0.044   0.9649    
---
Multiple R-squared:  0.7362,	Adjusted R-squared:  0.6922 
F-statistic: 16.74 on 7 and 42 DF,  p-value: 2.534e-10
```

## Example: Backward Elimination

```{r echo=TRUE, results='hide'}
lmod <- update(lmod, . ~ . - Area)
summary(lmod)
```

```
Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept)  7.099e+01  1.387e+00  51.165  < 2e-16 ***
Population   5.188e-05  2.879e-05   1.802   0.0785 .  
Income      -2.444e-05  2.343e-04  -0.104   0.9174    
Illiteracy   2.846e-02  3.416e-01   0.083   0.9340    
Murder      -3.018e-01  4.334e-02  -6.963 1.45e-08 ***
HS.Grad      4.847e-02  2.067e-02   2.345   0.0237 *  
Frost       -5.776e-03  2.970e-03  -1.945   0.0584 .  
```

## Example: Backward Elimination 

```{r echo=TRUE, results='hide'}
lmod <- update(lmod, . ~ . - Illiteracy)
summary(lmod)
```

```
Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept)  7.107e+01  1.029e+00  69.067  < 2e-16 ***
Population   5.115e-05  2.709e-05   1.888   0.0657 .  
Income      -2.477e-05  2.316e-04  -0.107   0.9153    
Murder      -3.000e-01  3.704e-02  -8.099 2.91e-10 ***
HS.Grad      4.776e-02  1.859e-02   2.569   0.0137 *  
Frost       -5.910e-03  2.468e-03  -2.395   0.0210 * 
```

## Example: Backward Elimination

```{r echo=TRUE, results='hide'}
lmod <- update(lmod, . ~ . - Income)
summary(lmod)
```

```
Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept)  7.103e+01  9.529e-01  74.542  < 2e-16 ***
Population   5.014e-05  2.512e-05   1.996  0.05201 .  
Murder      -3.001e-01  3.661e-02  -8.199 1.77e-10 ***
HS.Grad      4.658e-02  1.483e-02   3.142  0.00297 ** 
Frost       -5.943e-03  2.421e-03  -2.455  0.01802 * 
```

## Example: Backward Elimination

```{r echo=TRUE, results='hide'}
lmod <- update(lmod, . ~ . - Population)
summary(lmod)
```

```
Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept) 71.036379   0.983262  72.246  < 2e-16 ***
Murder      -0.283065   0.036731  -7.706 8.04e-10 ***
HS.Grad      0.049949   0.015201   3.286  0.00195 ** 
Frost       -0.006912   0.002447  -2.824  0.00699 ** 
---

Residual standard error: 0.7427 on 46 degrees of freedom
Multiple R-squared:  0.7127,	Adjusted R-squared:  0.6939 
F-statistic: 38.03 on 3 and 46 DF,  p-value: 1.634e-12
```

## Example: Backward Elimination

- Whether to remove `Population` is a close call as it is borderline.
- The $R^2$ value has changed only slightly, from $0.736$ to $0.713$.

## Problems with Stepwise Regression

1. No guarantee that we get the "best" possible subset of predictors.
2. Multiple testing.
    - Removing less significant predictors tends to increase the significance of those remaining.
    - May overstate actual significance of remaining predictors.
3. Procedure not directly linked to prediction or explanation, so may not be very helpful in our primary goal. 
4. Tends to pick models that are smaller than is ideal for prediction purposes. 

# Criterion-Based Procedures

## Criterion-Based Procedures

Idea: 

- Create some measure of how well a model meets some given purpose. 
- Select a model that best optimizes that criterion.

***

Natural to pick a model, $g$, parameterized by $\theta$, which is close to the true model $f$.

We can measure that distance using

$$I(f,g) = \int f(x)\log\left(\frac{f(x)}{g(x|\theta)}\right)dx$$

known as the *Kullback-Leibler information*. 

*** 

Since $f$ is unknown, we use the MLE of $\theta$

$$\hat{I}(f,g) = \int f(x)\log f(x)dx - \int f(x)\log g(x|\hat{\theta})dx$$

and we can ignore the first term since it doesn't depend on the model chosen.

## Akaike Information Criterion

$$AIC = -2\log L(\hat{\theta}) + 2p$$

where $p$ is the number of parameters in the model.

- Since this includes both model fit (likelihood) and complexity (number of parameters), it is a good measure for model parsimony.
    - $2p$ is considered a *penalty term*. 
- For linear regression, $$-2\log L(\hat{\theta}) = n\log(RSS/n) + \text{constant}$$
- We choose the model which minimized AIC. 

## Bayesian Informaton Criterion

$$BIC = -2\log L(\hat{\theta}) + p\log n$$

- Penalizes larger models more heavily than AIC does
- AIC usually preferred when prediction is primarly goal

## Example: AIC

```{r echo=TRUE}
require(leaps)
b <- regsubsets(Life.Exp~.,data=statedata)
rs <- summary(b)
rs$which
```

## Example: AIC

- The `regsubsets` function in the `leaps` package runs all possible combinations of the predictors.
- For each model size $p$, it finds the variables that produce the minimum RSS.

## Example: AIC

```{r}
rs$rss
AIC <- 50*log(rs$rss/50) + (2:8)*2
plot(AIC ~ I(1:7), ylab="AIC", xlab="Number of Predictors")
```

## Adjusted R-Squared

This is another commonly used criterion.

- As we increase $p$, $R^2$ is strictly non-decreasing. $$R^2 = 1 - \frac{\text{RSS}}{\text{TSS}}$$
- So $R^2$ is not an appropriate metric for model comparison. 
    
## Adjusted R-Squared
    
$$R^2_a =  1-\frac{\text{RSS}/(n-p)}{\text{TSS}/(n-1)} = 1-\left(\frac{n-1}{n-p}\right)(1-R^2)$$

- Adding a predictor will only increase $R^2_a$ if it has adequate predictive value.
- AIC and BIC can be used for any modelling approach, $R^2_a$ is specific to linear regression. 

## Example: Adjusted R-Squared

```{r}
rs$adjr2
plot(2:8,rs$adjr2,xlab="No. of Parameters",ylab="Adjusted R-square")
```

## Mallow's $C_p$

A good model should predict well, so we might examine the average MSE of prediction:

$$\frac{1}{\sigma^2}\sum_i E(\hat{y}_i - Ey_i)^2$$

which we can estimate using Mallow's $C_p$:

$$C_p = \frac{RSS_p}{\hat{\sigma}^2} + 2p - n$$

where $\hat{\sigma}^2$ is estimated from the model with all predictors and $RSS_p$ is the RSS from a model with $p$ parameters. 

## Mallow's $C_p$

- For the full model, $C_p = p$.
- If a $p$ predictor model fits, then $E(RSS) = (n-p)\sigma^2$ and $E(C_p) \approx p$.
- A model with a bad fit will have $C_p >> p$.

## Example: Mallow's $C_p$

```{r}
plot(2:8,rs$cp,xlab="No. of Parameters",ylab="Cp Statistic")
abline(0,1)
```

Models on or below the $C_p = p$ line are good fits. 

## Models with Many Predictors

- If there are $q$ potential predictors, there are $2^q$ possible models. 
- May not be practical to fit all possible models.
- The `step` function uses a search method to evaluate models sequentially using AIC

## Models with Many Predictors

```{r echo=TRUE, results='hide'}
lmod <- lm(Life.Exp ~ ., data=statedata)
step(lmod)
```

```
Start:  AIC=-22.18
Life.Exp ~ Population + Income + Illiteracy + Murder + HS.Grad + 
    Frost + Area

             Df Sum of Sq    RSS     AIC
- Area        1    0.0011 23.298 -24.182
- Income      1    0.0044 23.302 -24.175
- Illiteracy  1    0.0047 23.302 -24.174
<none>                    23.297 -22.185
- Population  1    1.7472 25.044 -20.569
- Frost       1    1.8466 25.144 -20.371
- HS.Grad     1    2.4413 25.738 -19.202
- Murder      1   23.1411 46.438  10.305
```

*** 

```
Step:  AIC=-24.18
Life.Exp ~ Population + Income + Illiteracy + Murder + HS.Grad + 
    Frost

             Df Sum of Sq    RSS     AIC
- Illiteracy  1    0.0038 23.302 -26.174
- Income      1    0.0059 23.304 -26.170
<none>                    23.298 -24.182
- Population  1    1.7599 25.058 -22.541
- Frost       1    2.0488 25.347 -21.968
- HS.Grad     1    2.9804 26.279 -20.163
- Murder      1   26.2721 49.570  11.569
```

***

```
Step:  AIC=-26.17
Life.Exp ~ Population + Income + Murder + HS.Grad + Frost

             Df Sum of Sq    RSS     AIC
- Income      1     0.006 23.308 -28.161
<none>                    23.302 -26.174
- Population  1     1.887 25.189 -24.280
- Frost       1     3.037 26.339 -22.048
- HS.Grad     1     3.495 26.797 -21.187
- Murder      1    34.739 58.041  17.456
```

***

```
Step:  AIC=-28.16
Life.Exp ~ Population + Murder + HS.Grad + Frost

             Df Sum of Sq    RSS     AIC
<none>                    23.308 -28.161
- Population  1     2.064 25.372 -25.920
- Frost       1     3.122 26.430 -23.877
- HS.Grad     1     5.112 28.420 -20.246
- Murder      1    34.816 58.124  15.528

Call:
lm(formula = Life.Exp ~ Population + Murder + HS.Grad + Frost, 
    data = statedata)

Coefficients:
(Intercept)   Population       Murder      HS.Grad        Frost  
  7.103e+01    5.014e-05   -3.001e-01    4.658e-02   -5.943e-03  
```

## Other Concerns

- Variable selection methods are sensitive to outliers and high leverage points.
    - We should examine these and examine how they change if these points are removed.
- Transformations may also have an impact.