---
title: "hw5"
author: "Austin Melendez"
date: "2025-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### The following was completed for the STAT 140A - Homework 5 assignment.

# Homework 5

## Using the sat dataset, fit a model with the total SAT score as the response and expend, salary, ratio and takers as predictors. 

### 6.1) Perform regression diagnostics on this model to answer the following questions. Display any plots that are relevant.

```{r}
#install.packages("faraway")
library(faraway)

data(sat)

model <- lm(total ~ expend + salary + ratio + takers, data = sat)

summary(model)
```

#### A) Lets check our model assumptions to verify the model is useful. First, let's check the constant variance assumption, we'll test for homoscedasticity.

```{r}
plot(model, which = 1)

plot(model, which = 3)
```

The data shows some possible mild heteroscedasticity, but not a large influence. The residuals vs fitted graph shows some points we may want to investigate further such as `West Virginia`, `North Dakota`, and `New Hampshire`.

#### B) Next lets check the normality assumption.

```{r}
plot(model, which = 2)
```

Looking at the Q-Q plot, some the majority of our data is on the line. However, it could be argued the the extreme ends of the data ends with tailing values. Based on this graph, `West virginia` is definitely a point to investigate further.

#### C) Lets take a closer look at some high leverage points.

```{r}
hatv <- hatvalues(model)

print(hatv)

sum(hatv)
```

Two high leverage points that I can identify on on this graph are `California` and `Utah`.

#### D) Let's check for outliers!

```{r}

plot(model, which = 4)

which(abs(rstandard(model)) > 2)
```

This shows us that there are four states that exceed $\pm2$ standard deviations. These four states are `New Hampshire`, `North Dakota`, `Utah`, and `West Virginia`.

#### E) Influential points?

```{r}
states <- row.names(sat)

cook <- cooks.distance(model)

halfnorm(cook,3,labs=states,ylab="Cook’s distances")
```

`West Virginia`, `New Hampshire`, and `Utah` have high cook's distances with high half-normal quantiles. So let's remove the highest value, `Utah`, from the model to see if the model improves.

#### F) Model Structure?

```{r}
model_reduced <- lm(total ~ expend + salary + ratio + takers, 
                    data = subset(sat, rownames(sat) != "Utah"))

summary(model_reduced)
```

### 6.2) sing the teengamb dataset answer the following questions.

```{r}
data(teengamb)

model2 <- lm(gamble ~ sex + status + income + verbal, data = teengamb)

summary(model2)
```

#### A) Lets check our model assumptions to verify the model is useful. First, let's check the constant variance assumption, we'll test for homoscedasticity.

```{r}
plot(model2, which = 1)

plot(model2, which = 3)
```

The data definitely shows heteroscedasticity and possibly non-linearity. The residuals vs fitted graph shows an obvious cone of points that increases distances as you go through the data.

#### B) Next lets check the normality assumption.

```{r}
plot(model2, which = 2)
```

Looking at the Q-Q plot, some the majority of our data is on the line. However, the the extreme ends of the data ends with tailing values. Based on this graph, points `24`, `36`, and `39` are definitely points to investigate further.

#### C) Lets take a closer look at some high leverage points.

```{r}
hatv2 <- hatvalues(model2)

print(hatv2)

sum(hatv2)
```

Two high leverage points that I can identify on on this graph are `24` and `39`.

#### D) Let's check for outliers!

```{r}

plot(model2, which = 4)

which(abs(rstandard(model2)) > 2)
```

This shows us that there are three points that exceed $\pm2$ standard deviations. These three points are `24`, `36`, and `39`.

#### E) Influential points?

```{r}
index <- row.names(teengamb)

cook2 <- cooks.distance(model2)

halfnorm(cook,3,labs=index,ylab="Cook’s distances")
```

`24` and `39` have high cook's distances with high half-normal quantiles. So let's remove the highest value, `24`, from the model to see if the model improves.

#### F) Model Structure?

```{r}
model_reduced2 <- lm(gamble ~ sex + status + income + verbal, data = subset(teengamb, rownames(teengamb) != "24"))

summary(model_reduced2)
```

### 7.3) Using the divusa data:

#### A) it a regression model with divorce as the response and unemployed, femlab, marriage, birth and military as predictors.

```{r}
data(divusa)

model3 <- lm(divorce ~ unemployed + femlab + marriage + birth + military, data = divusa)
summary(model3)

kappa(model3, exact = TRUE)
```

A high condition number indicates that some predictors are highly linearly related coefficients may be unstable and standard errors inflated.

#### B) Compute VIFs

```{r}
vif(model3)
```

The VIF's show us there is low multicollinerarity.

#### C) Remove insignificant predictors and check collinerarity.

```{r}
model3_reduced <- lm(divorce ~ unemployed + femlab + marriage, data = divusa)

summary(model3_reduced)

vif(model3_reduced)

kappa(model3_reduced, exact = TRUE)
```

The VIF's have decreased so the collinearity has be en reduced.

### 9.2) he cornnit data on the relationship between corn `yield` (bushels per acre) and `nitrogen` (pounds per acre) fertilizer application.

```{r}
data(cornnit)

model4 <- lm(yield ~ nitrogen, data = cornnit)

summary(model4)

plot(model4, which = 1)
```

This shows a slight non-linear pattern that we will have to investigate further.

Let's try some different common transformations.

```{r}
model_log <- lm(log(yield) ~ nitrogen, data=cornnit)
plot(model_log$fitted.values, resid(model_log))

abline(h=0, lty=2)

model_sqrt <- lm(sqrt(yield) ~ nitrogen, data=cornnit)

model_quad <- lm(yield ~ nitrogen + I(nitrogen^2), data=cornnit)

summary(model_quad)
```

Let's compare the goodness of fit for our transformed models.

```{r}
anova(model4, model_quad)
summary(model_quad)
```
Lastly, lets do a Box-Cox to automatically find the optimal transformation.

```{r}
library(MASS)
boxcox(model4, lambda = seq(-2, 2, 0.1))
```

Box-Cox suggests a lambda of $\lambda \ 1$ which suggests using a log transformation.

### 9.3 Using the ozone data, fit a model with `O3` as the response and `temp`, `humidity` and `ibh` as predictors. Use the Box–Cox method to determine the best transformation on the response.

```{r}
data(ozone)

model_oz <- lm(O3 ~ temp + humidity + ibh, data=ozone)

summary(model_oz)

plot(model_oz, which = 1)
```

The residuals vs fitted plot indicates non-constant variance, we will need to transform this data to fix it.

```{r}
boxcox(model_oz, lambda = seq(-2, 2, 0.1))
```

Box-Cox suggests a lambda of $\lambda \approx 0$ which suggests using a log transformation.

```{r}
model_oz_log <- lm(log(O3) ~ temp + humidity + ibh, data=ozone)

summary(model_oz_log)
```

