---
title: "hw6"
author: "Austin Melendez"
date: "2025-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1
Use the following three lines of code to generate data for a linear regression.
```{r}
library(tidyverse)
library(faraway)
data(infmort)

summary(infmort)

mod <- lm(mortality ~ ., data=infmort)
summary(mod)
```
From this model using all other predictor values to predict `mortality`, all of our predictors are statistically significant besides `income`. However, our $R^2$ value is very low, 0.3105, which means it is only capturing ~31% of the data. Lets see if we can make that better.

Lets use AIC to optimize the model complexity vs significance.

```{r}
mod_step <- step(mod, k=2)
summary(mod_step)
```

The AIC slightly lowers when removing the `income` estimator. Let's check the Box-Cox plot to verify our model assumptions.

```{r}
library(MASS)
bc <- boxcox(mod_step, plotit=TRUE)
```

Since the Box-Cox suggests lambda ~ 0, so we transform the response variable accordingly with a log transformation.
```{r}
mod_trans <- lm(log(mortality) ~ region + oil, data=infmort)
summary(mod_trans)
```
THis has almost doubled our $R^2$ value. Next, lets check for influential points.

```{r}
plot(mod_step, 2)

plot(mod_step, 4)

cooks <- cooks.distance(mod_step)
```
Based on our Q-Q plot I would like to investiage `Saudi_Arabia` and `Afganistan` as they have significantly higher residuals. We can see on our Residuals vs Leverage plot that `Saudi_Arabia` is definitely a high influential point that needs to be address as it is inside the Cook's distance dashed arc.

Let's see which points have high leverage.

```{r}
hat <- hatvalues(mod_step)
which(hat > 2*mean(hat))
```
Let's refit our model without the highly influential point, `Saudi Arabia`.

```{r}
infmort2 <- subset(infmort, row.names(infmort) != "Saudi_Arabia")
mod_reduced <- lm(log(mortality) ~ region + oil, data = infmort2)

summary(mod_reduced)
```

This is our final model.

## 2
Let's first convert the predictor variable `side` to factor.
```{r}
data(chredlin)
summary(chredlin)

```
Let's fit a model with side as a factor and fit the model with interactor to allow the slope to vary by side.

```{r}
mod_side <- lm(race ~ fire + theft + age + involact + income + side, 
               data = chredlin)
summary(mod_side)

mod_side_int <- lm(race ~ (fire + theft + age + involact + income) * side, 
                   data = chredlin)
summary(mod_side_int)
```

Now let's compare these two models.

```{r}
anova(mod_side, mod_side_int)

plot(mod_side_int)
```

## 6 
### A)
```{r}
data(clot)
summary(clot)

clot$lot <- factor(clot$lot)  # ensure factor
plot(clot$conc, clot$time, pch=ifelse(clot$lot==levels(clot$lot)[1],1,2),
     xlab="conc (%)", ylab="time")
legend("topright", legend=levels(clot$lot), pch=1:2)

```
### B) 
Let's check out our response and predictor varibales to see if we need to perfrom any transformations.

```{r}
mod <- lm(time ~ conc * lot, data = clot)
boxcox(mod, plotit = TRUE)
```

As we can see in our Box-Cox plot, the $\lambda$ value is around $-1.5$. 
### C)
This means we should do a log log transformation on our response variable. 

```{r}
mod_loglog <- lm(log(time) ~ log(conc) * lot, data = clot)
summary(mod_loglog)
```

Let's also compare our models using an ANOVA table.

```{r}
anova(lm(log(time) ~ log(conc) + lot, data=clot), mod_loglog)
```
### D)

```{r}
shapiro.test(resid(mod_loglog))
plot(fitted(mod_loglog), resid(mod_loglog)); abline(h=0)
```

## 8
### A
```{r}
data(hips)
summary(hips)

plot(hips$fbef, hips$fafter, pch=ifelse(hips$grp=="control",1,2),
     xlab="flexion before", ylab="flexion after")
legend("bottomright", legend=c("control","treat"), pch=1:2)
```

### B)
```{r}
mod_ancova <- lm(faft ~ fbef + grp, data = hips)
summary(mod_ancova)
anova(mod_ancova)
```

Let;s check if the slopes are different by treatment (`grp`). 
```{r}
mod_anc_int <- lm(faft ~ fbef * grp, data = hips)
anova(mod_ancova, mod_anc_int)
```

### C)
```{r}
hips$diff <- hips$faft - hips$fbef
boxplot(diff ~ grp, data = hips)

t.test(diff ~ grp, data = hips, var.equal = FALSE)
```

ANCOVA adjusts `fafter` for baseline and typically has better power if baseline correlates with outcome. While the difference approach directly tests change.

### D)
Let's check for out highly influential points.

```{r}
which(hips$fbef < 90)

plot(mod_ancova, 2)
plot(mod_ancova, 4)
```

We see points `49` and `50` are both very high Cook's distance and high leverage. Let see which other points are considered high.

```{r}
cooks <- cooks.distance(mod_ancova)
which(cooks > 4/length(cooks))
```

So, let's remove the outliers that are high leverage and highly influential and refit our model.

```{r}
hips2 <- hips[-c(49, 50), ]

mod_trim <- lm(faft ~ fbef + grp, data = hips2)
summary(mod_trim)
```

### E)

```{r}
confint(mod_trim)["grptreat",]
```

### G)

```{r}
hips_avg <- hips %>%
  group_by(person, grp) %>%
  summarize(
    fbef_mean = mean(fbef),
    faft_mean = mean(faft),
    diff_mean = mean(diff),
    .groups = "drop"
  )

mod_avg <- lm(faft_mean ~ fbef_mean + grp, data = hips_avg)

summary(mod_avg)
confint(mod_avg)
```



