---
title: "hw7"
author: "Austin Melendez"
date: "2025-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1
Use the `prostate` data wiht `lspa` as the response and other variables as predictors. Determine the "best" model:

```{r}
#install.packages("faraway")
library(faraway)
data(prostate)
head(prostate)

full_mod <- lm(lpsa ~ ., data = prostate)
summary(full_mod)
```

### A
#### Backward eliminination

```{r}
back_mod <- step(full_mod, direction = "backward")
summary(back_mod)
```

According to the backward elimination, the strongest predictors of `lspa` are `lcavol`, `lweight`, `svi`, and `lbph`.

### B
#### AIC Selection

```{r}
library(MASS)
aic_mod <- stepAIC(full_mod, direction = "both")
summary(aic_mod)
```

According to the AIC selection process, the strongest predictors of `lspa` are `lcavol`, `lweight`, `lbph`, and `svi`.

### C
#### Adjusted R^2

```{r}
#install.packages("leaps") 
require(leaps)
adjr2_fit <- regsubsets(lpsa ~ ., data = prostate)
adjr2_sum <- summary(adjr2_fit)

adjr2_sum$which[which.max(adjr2_sum$adjr2), ]
```
The model with the highest adjusted $ R^2 $ uses the predictors of `lcavol`, `lweight`, `age`, `lbph`, `svi`, `lcp`, and `pgg45`.

### D
#### Mallows Cp

```{r}
cp <- adjr2_sum$cp
p <- 1:length(cp)
adjr2_sum$which[which.min(abs(cp - p)), ]
```
Mallow's Cp confirms best model uses the predictors of `lcavol`, `lweight`, `lbph`, `svi`, `lcp`, and `pgg45`.

## 2
Use `gamble` as the response and determine the best subset of predictors.

```{r}
library(faraway)
data(teengamb)
head(teengamb)

full_tg <- lm(gamble ~ ., data = teengamb)
summary(full_tg)
```

### A
#### Backward Elimination

```{r}
backward_tg <- step(full_tg, direction = "backward")
summary(backward_tg)
```

Backward elimination shows that the best mnodel uses the predictors of `sex`, `income`, and `verbal`.

### B
#### AIC

```{r}
library(MASS)
aic_tg <- stepAIC(full_tg, direction = "both")
summary(aic_tg)
```
AIC shows that the best mnodel uses the predictors of `sex`, `income`, and `verbal`.

### C
#### Adjusted R^2

```{r}
require(leaps)
adjr2_tg <- regsubsets(gamble ~ ., data = teengamb)
adjr2_sum_tg <- summary(adjr2_tg)
which.max(adjr2_sum_tg$adjr2)
adjr2_sum_tg$which[which.max(adjr2_sum_tg$adjr2), ]
```
Adjusted $ R^2 $ shows the best model uses the predictors of `sex`, `income`, and `verbal`.

### D
#### Mallows Cp

```{r}
cp_tg <- adjr2_sum_tg$cp
p_tg <- 1:length(cp_tg)
adjr2_sum_tg$which[which.min(abs(cp_tg - p_tg)), ]
```
Mallows Cp shows the same best model uses the predictors of `sex`, `income`, and `verbal`.

## 1
Determine whether there are any differences between the operators in the `pulp` dataset.

```{r}
data(pulp)
head(pulp)

lm_pulp <- lm(bright ~ operator, data = pulp)
anova(lm_pulp)
summary(lm_pulp)
```
```{r}
plot(lm_pulp)
```
After viewing the plots, the variance seems to be constant among groups, and the Q-Q plot looks fine.

```{r}
TukeyHSD(aov(lm_pulp))
```
Looking at our table of values, we only have one significant difference and one difference that is borderline significant at best.

Operator D produces significantly higher mean brightness than operator B.

The other pairwise differences are not significant, suggesting only slight operator effects on brightness levels.

## 2
Determine whether there are any differences between the operators in the `chickwts` dataset.

```{r}
data(chickwts)
lm_chick <- lm(weight ~ feed, data = chickwts)
anova(lm_chick)
summary(lm_chick)
```

```{r}
plot(lm_chick)
```
After viewing the plots, the variance seems to be constant among groups, and the Q-Q plot looks fine.

```{r}
TukeyHSD(aov(lm_chick))
```
Looking at our table of values, we have a few significant differences. We notice a difference in `horsebean` and `casein`, `linseed` and `casin`, `soybean` and `casein`, `meatmeal` and `horsebean`, `soybean` and `horsebean`, `sunflower` and `linseed`, and lastly `sunflower` and `soybean`.

The other pairwise differences are not significant, however the amount of significant differences suggests significant differences in mean chick weights based on feed type. 

`Casein` feed produces significantly heavier chicks than `horsebean`, `linseed`, and `soybean` feeds.

`Horsebean` feed has the lowest average weight among all feed types.

`Sunflower` feed produces the highest mean weight.

## 3
Determine whether there are any differences between the operators in the `plantgrowth` dataset.

```{r}
data(PlantGrowth)
lm_pg <- lm(weight ~ group, data = PlantGrowth)
anova(lm_pg)
summary(lm_pg)
```

```{r}
plot(lm_pg)
```
After viewing the plots, the variance seems to be constant among groups, and the Q-Q plot looks fine.

```{r}
TukeyHSD(aov(lm_pg))
```
Looking at our table of values, we notice the only significant difference is between `trt2` and `trt1`.

No significant difference between the control and either treatment individually. However, `trt2` has a significantly higher mean plant weight than `trt1`.

```{r}
contrasts(PlantGrowth$group) <- matrix(c(
  -2,  0,   # ctrl
   1,  1,   # trt1
   1, -1    # trt2
), ncol = 2)

lm_pg2 <- lm(weight ~ group, data = PlantGrowth)
summary(lm_pg2)
```
`trt2` plants are about 0.53 units lighter than `ctrl`.

There are significant differences in mean plant weight between the three treatment groups.
