---
title: "ISLR2 Exam Cheat Sheet: Chapters 6-9"
author: "ISLR2 Lab Summary"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Required Libraries
library(ISLR2)
library(leaps)    # Chapter 6: Selection
library(glmnet)   # Chapter 6: Ridge/Lasso
library(pls)      # Chapter 6: PCR/PLS
library(splines)  # Chapter 7: Splines
library(gam)      # Chapter 7: GAMs
library(tree)     # Chapter 8: Trees
library(randomForest) # Chapter 8: Bagging/RF
library(gbm)      # Chapter 8: Boosting
library(BART)     # Chapter 8: BART
library(e1071)    # Chapter 9: SVM
```

# Chapter 6: Linear Model Selection & Regularization
## Best Subset and Stepwise Selection
We use regsubsets() to find the best model for each size. By default, it only goes up to 8 variables; use nvmax to increase this.

```{r}
# Data cleaning for Hitters dataset
data("Hitters")
Hitters <- na.omit(Hitters)

# Best Subset Selection
regfit.full <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19)
reg.summary <- summary(regfit.full)

# Plotting metrics: Adjusted R2, Cp, and BIC
par(mfrow = c(2, 2))
plot(reg.summary$rss, xlab = "Variables", ylab = "RSS", type = "l")
plot(reg.summary$adjr2, xlab = "Variables", ylab = "Adjusted R2", type = "l")
points(which.max(reg.summary$adjr2), reg.summary$adjr2[which.max(reg.summary$adjr2)], col = "red", cex = 2, pch = 20)

# Forward/Backward Stepwise
regfit.fwd <- regsubsets(Salary ~ ., data = Hitters, nvmax = 19, method = "forward")
```

## Ridge Regression and Lasso
Requires a matrix input x and a vector y. Use cv.glmnet for automatic selection of the tuning parameter $\lambda$.
```{r}
x <- model.matrix(Salary ~ ., Hitters)[, -1]
y <- Hitters$Salary

# alpha = 0 is Ridge, alpha = 1 is Lasso
cv.out <- cv.glmnet(x, y, alpha = 1)
bestlam <- cv.out$lambda.min

# Final model coefficients
lasso.coef <- predict(cv.out, type = "coefficients", s = bestlam)
lasso.coef
```

# Chapter 7: Moving Beyond Linearity
## Polynomials and Splines
poly() creates orthogonal polynomials by default. Splines allow for localized fitting using knots.

```{r}
# Polynomial Degree 4
fit_poly <- lm(wage ~ poly(age, 4), data = Wage)

# Splines: bs() for B-splines, ns() for Natural Splines (stable at boundaries)
fit_spline <- lm(wage ~ ns(age, df = 4), data = Wage)

# Smoothing Spline: Automatic DOF via Generalized Cross-Validation
fit_smooth <- smooth.spline(Wage$age, Wage$wage, cv = TRUE)
```

## Generalized Additive Models (GAMs)
GAMs use the gam() function to fit additive models with smoothing terms s().

```{r}
gam_m3 <- gam(wage ~ s(year, 4) + s(age, 5) + education, data = Wage)
par(mfrow = c(1, 3))
plot(gam_m3, se = TRUE, col = "blue")
```

# Chapter 8: Tree-Based Methods
## Classification and Regression Trees
Trees are grown via tree() and then pruned via cost-complexity pruning using cv.tree().
```{r}
# Create binary classification variable
High <- factor(ifelse(Carseats$Sales <= 8, "No", "Yes"))
Carseats_df <- data.frame(Carseats, High)

# Fit and Prune
tree_carseats <- tree(High ~ . - Sales, Carseats_df)
cv_carseats <- cv.tree(tree_carseats, FUN = prune.misclass)
# Select best tree size (e.g., 9 nodes)
pruned_tree <- prune.misclass(tree_carseats, best = 9)

plot(pruned_tree)
text(pruned_tree, pretty = 0)
```

```{r}
data(Carseats)

set.seed(1)
train <- sample(1:nrow(Carseats), nrow(Carseats)/2)

tree_fit <- tree(Sales ~ ., data = Carseats, subset = train)

rf_fit <- randomForest(Sales ~ ., data = Carseats, subset = train)
rf_fit
importance(rf_fit)
```


## Random Forests, Boosting, and BART
Ensemble methods combine multiple trees to improve accuracy.
```{r}
# Random Forest: mtry < p (usually sqrt(p) for class, p/3 for reg)
rf_boston <- randomForest(medv ~ ., data = Boston, mtry = 6, importance = TRUE)

# Boosting: Sequential growth. shrinkage = learning rate
boost_boston <- gbm(medv ~ ., data = Boston, distribution = "gaussian", 
                    n.trees = 5000, interaction.depth = 4, shrinkage = 0.01)

# BART: Bayesian approach
# bart_fit <- gbart(x_train, y_train, x.test = x_test)
```

# Chapter 9: Support Vector Machines
## Support Vector Classifier (Linear Kernel)
Used for classification. The dependent variable y must be a factor.

```{r}
# kernel = "linear" creates the Support Vector Classifier
svm_linear <- svm(High ~ Price + Income, data = Carseats_df, 
                  kernel = "linear", cost = 10)
summary(svm_linear)
```

## Support Vector Machine (Radial/Polynomial)
Use tune() to perform cross-validation over a grid of cost and gamma parameters.
```{r}
# Radial Kernel
set.seed(1)
tune_out <- tune(svm, High ~ Price + Income, data = Carseats_df, kernel = "radial",
                 ranges = list(cost = c(0.1, 1, 10), gamma = c(0.5, 1, 2)))

# Best model
best_svm <- tune_out$best.model
plot(best_svm, Carseats_df, Price ~ Income)
```

