---
title: "Modeling Reference Sheet"
author: "Austin Melendez"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
#install.packages("MASS")
#install.packages("e1071")
#install.packages("ISLR2")
#install.packages("class")
#install.packages("boot")
#install.packages("leaps")
#install.packages("glmnet")
#install.packages("pls")
```


# STAT140B - Statistical Modeling
## Modeling Reference Cheat Sheet
### Austin Melendez
#### 3/5/2026

```{r}
# Load required libraries
library(MASS)   # Boston dataset, lda, qda
library(ISLR2)  # Default dataset
library(e1071)  # Naive Bayes
library(class)  # KNN
library(boot)   # CV and Bootstrap
library(leaps)  # Subset selection
library(glmnet) # Ridge and Lasso
library(pls)    # PCR and PLS
```


# Chapter 2 - Simple Linear Regression
* Predict quantitative Y using one predictor

```{r}
# Using Boston dataset: medv (Value) ~ lstat (Lower Status %)
fit <- lm(medv ~ lstat, data = Boston)

# Summary
summary(fit)

# Confidence intervals for coefficients
confint(fit)

# Predictions for a specific lstat value (e.g., 5%)
predict(fit, newdata = data.frame(lstat = 5), interval = "confidence")
predict(fit, newdata = data.frame(lstat = 5), interval = "prediction")
```

# Chapter 3 - Multiple Linear Regression
* Predict quantitative outputs using many predictors
* Check variable importance
* Can evaluate interactions and nonlinear terms
```{r}
# Predict medv using lstat, age, and rm (average rooms)
fit <- lm(medv ~ lstat + age + rm, data = Boston)
summary(fit)

# Interaction between lstat and age
fit.int <- lm(medv ~ lstat * age, data = Boston)

# Polynomial regression (degree 3)
fit.poly <- lm(medv ~ poly(lstat, 3), data = Boston)

# Diagnostics
par(mfrow = c(2, 2))
plot(fit)
```

# Chapter 4 - Classification Methods
```{r}
# Prepare data
data(Default)
set.seed(1)
train_idx <- sample(1:nrow(Default), 0.7 * nrow(Default))
train_data <- Default[train_idx, ]
test_data  <- Default[-train_idx, ]
```

### Logistic Regression
* Binary classifications
* Produces probabilities of landing in a class
* Interpretabble coefficients (log odds)
```{r}
log.fit <- glm(default ~ balance + income, data = train_data, family = binomial)
summary(log.fit)

probs <- predict(log.fit, test_data, type = "response")
preds <- ifelse(probs > 0.5, "Yes", "No")

# Accuracy
mean(preds == test_data$default)
```

### Linear Discriminant Analysis (LDA)
* Classification when classes are well seperated
* Assumes predicators are normally dist
* Works better than logistic regression when $n$ is small
* Works when deicision boundaries are linear
```{r}
lda.fit <- lda(default ~ balance + income, data = train_data)
lda.pred <- predict(lda.fit, test_data)$class

mean(lda.pred == test_data$default)
```

### Quadratic Discriminant Analysis (QDA)
* Like LDA but allows each class its own covariance
* Flexible, works when decision boundaries are nonlinear
```{r}
qda.fit <- qda(default ~ balance + income, data = train_data)
qda.pred <- predict(qda.fit, test_data)$class

mean(qda.pred == test_data$default)
```

### Naive Bayes
* Classifier assuming independece of predictors
* Works well with many predictors
* Good for text data
```{r}
nb.fit <- naiveBayes(default ~ balance + income, data = train_data)
nb.pred <- predict(nb.fit, test_data)

mean(nb.pred == test_data$default)
```

### K-Nearest Neighbors (KNN)
* Nonparametric classification
* Good when decision boundary is irregular
* Must scale/normalize data!
```{r}
# Scale numeric predictors for KNN
train.X <- scale(train_data[, c("balance", "income")])
test.X  <- scale(test_data[, c("balance", "income")])
train.Y <- train_data$default

knn.pred <- knn(train.X, test.X, train.Y, k = 5)

mean(knn.pred == test_data$default)
```

# Chapter 5 - Resampling Methods
### Validation Set Approach
* Split into train and test set
* Only trains on 50-70% of data (high variance)
* The model is trained on the training set and evaluated on the validation set.
```{r}
set.seed(1)
train.idx <- sample(1:nrow(Boston), nrow(Boston)/2)
train <- Boston[train.idx, ]
test  <- Boston[-train.idx, ]

fit <- lm(medv ~ ., data = train)
pred <- predict(fit, newdata = test)
mean((test$medv - pred)^2) # Test MSE
```

### Leave-One-Out Cross Validation (LOOCV)
* A special case of k-fold cross-validation where $k = n$
* Each observation is left out one at a time. Train model on $n−1$ observations, then test on the one left out observation. Average the prediction errors
* Use with small data sets, low bias and possible high variance
```{r}
# cv.glm works with glm objects
fit.glm <- glm(medv ~ lstat + age, data = Boston)
loocv.err <- cv.glm(Boston, fit.glm)

loocv.err$delta[1] # Raw CV error
```

### k-Fold Cross Validation
* The data is divided into $k$ equal-sized folds
* Split data into $k$ groups and train model on $k-1$ folds. Test on remaining folds and repeat until each fold has been used as the validation set. Then average the errors.
* Typical $k = 5$ or $k = 10$
* Use with moderate data sets, good balance between bias and variance
```{r}
set.seed(1)
cv.err <- cv.glm(Boston, fit.glm, K = 10)
cv.err$delta[1]
```

### Bootstrap
* Bootstrap repeatedly samples observations with replacement from the dataset to create many simulated datasets called bootstrap samples
* Sample n observations with replacement, then fit model and compute statistic of interest (mean, median, CI, errors, bias, etc). Repeat process many times.
```{r}
boot.fn <- function(data, index) {
  coef(lm(medv ~ lstat + age, data = data, subset = index))
}

boot(Boston, boot.fn, R = 1000)
```

# Chapter 6 - Model Selection and Regularization
## Select Subset of Predictors Methods
### Best Subset Selection
* Best subset selection fits all possible models using every combination of predictors and selects the best model for each model size
* Use when the number of predictors p is small
* Uses Cp, BIC, adj-$R^2$
```{r}
# Using Boston data to predict medv
regfit <- regsubsets(medv ~ ., data = Boston, nvmax = 13)
reg.summary <- summary(regfit)

par(mfrow = c(1, 3))
plot(reg.summary$bic, type = "l", main = "BIC")
plot(reg.summary$cp, type = "l", main = "Cp")
plot(reg.summary$adjr2, type = "l", main = "Adj R2")

# Identify the best model according to BIC
which.min(reg.summary$bic)
```

### Forward / Backward Stepwise Selection
* Forward selection starts with no predictors and adds predictors one at a time
* Use forward selection when p is moderately large and works even when $p > n$
* Backward selection starts with the full model containing all predictors and removes the least useful predictor at each step
* Use backward selection when p is moderate and works only when $n > p$
* Both may miss the optimal model, depends on the order variables are added
```{r}
reg.fwd <- regsubsets(medv ~ ., data = Boston, nvmax = 13, method="forward")

reg.bwd <- regsubsets(medv ~ ., data = Boston, nvmax = 13, method="backward")
```

## Shrink Coefficient Methods
### Ridge Regression
* Ridge regression modifies least squares by adding a penalty on the squared size of coefficients
* Use with many predictors and possible multicollinearity
* You want coefficients to shrink towards zero but not be removed
* Use when many predictors exist and they are highly correlated. We can reduce variance without removing predictors
* Reduces overfitting and handles multicollinearity well
```{r}
# Prepare matrix input
X <- model.matrix(medv ~ ., data = Boston)[, -1]
y <- Boston$medv

# Ridge (alpha = 0)
cv.ridge <- cv.glmnet(X, y, alpha = 0)
plot(cv.ridge)

# Coefficients for best Lasso model
best.lambda <- cv.ridge$lambda.min
predict(cv.ridge, s = best.lambda, type = "coefficients")
```

### Lasso
* Lasso adds a penalty based on the absolute value of coefficients, this penalty can shrink some coefficients exactly to zero
* Use when you believe only a subset of predictors are important
* Can behave unpredictably when predictors are highly correlated
```{r}
# Lasso (alpha = 1)
cv.lasso <- cv.glmnet(X, y, alpha = 1)
plot(cv.lasso)

# Coefficients for best Lasso model
best.lambda <- cv.lasso$lambda.min
predict(cv.lasso, s = best.lambda, type = "coefficients")
coef(cv.lasso, s=best.lambda)
```

## Dimension Reduction Methods
### Principal Components Regression (PCR)
* PCR first performs principal component analysis (PCA) on the predictors and then performs regression using the principal components instead of the original variables
* The components capture the directions of greatest variance in the predictors
* Use PCR when predictors are highly correlated, p is large, and we want to reduce dimensionality before regression
```{r}
# PCR
pcr.fit <- pcr(medv ~ ., data = Boston, scale = TRUE, validation = "CV")
validationplot(pcr.fit, val.type = "MSEP")
```

### Partial Least Squares (PLS)
* PLS is similar to PCR but differs because it chooses components using both predictors and the response variable
* This means the components are selected to maximize covariance with the response
* Use PLS when the predictors are highly correlated, you want dimension reduction, and you want components that directly help predict Y
```{r}
# PLS
pls.fit <- plsr(medv ~ ., data = Boston, scale = TRUE, validation = "CV")
summary(pls.fit)
```

# Chapter 7 - Moving Beyond Linearity
### Polynomial Regression and Step Functions
* Polynomials: Extending linear models by adding extra predictors (powers of X).
* Step Functions: Breaking the range of X into bins.
```{r}
# Polynomial Regression (Degree 4)
fit.poly <- lm(medv ~ poly(lstat, 4), data = Boston)

# Step Functions
table(cut(Boston$lstat, 4)) # Create 4 bins
fit.step <- lm(medv ~ cut(lstat, 4), data = Boston)
```

### Splines and GAMs
* Regression Splines: Piecewise polynomials joined at "knots."
* Smoothing Splines: Minimizes RSS with a roughness penalty.
* GAMs (Generalized Additive Models): Allows for non-linear functions of multiple predictors.
```{r}
library(splines)
library(gam)

# Regression Spline with specified knots
fit.spline <- lm(medv ~ bs(lstat, knots = c(10, 20, 30)), data = Boston)

# GAM using Smoothing Splines (s)
fit.gam <- gam(medv ~ s(lstat, 4) + s(rm, 4), data = Boston)
par(mfrow = c(1, 2))
plot(fit.gam, se = TRUE, col = "blue")
```

# Chapter 8 - Tree-Based Methods
### Decision Trees
Simple, interpretable, but prone to overfitting.
```{r}
library(tree)

# Classification Tree
tree.default <- tree(default ~ ., data = train_data)
plot(tree.default)
text(tree.default, pretty = 0)

# Pruning
cv.default <- cv.tree(tree.default, FUN = prune.misclass)
prune.default <- prune.misclass(tree.default, best = 5)
```

### Bagging and Random Forests
* Bagging: Building multiple trees from bootstrapped samples and averaging them.
* Random Forest: Like bagging, but decorrelates trees by only considering a subset of predictors at each split.
```{r}
library(randomForest)

# Bagging (mtry = total number of predictors)
bag.boston <- randomForest(medv ~ ., data = Boston, mtry = 13, importance = TRUE)

# Random Forest (mtry = default is p/3 for regression, sqrt(p) for class)
rf.boston <- randomForest(medv ~ ., data = Boston, importance = TRUE)
importance(rf.boston)
varImpPlot(rf.boston)
```

### Boosting
Builds trees sequentially; each tree learns from the residuals of the previous ones.
```{r}
library(gbm)

# Boosting (Distribution "gaussian" for regression, "bernoulli" for binary class)
boost.boston <- gbm(medv ~ ., data = Boston, distribution = "gaussian", 
                    n.trees = 5000, interaction.depth = 4)

summary(boost.boston) # Variable importance
```

# Chapter 9 - Support Vector Machines
### Support Vector Classifier
Linear boundary: uses a "cost" parameter to allow for observations to be on the wrong side of the margin.
```{r}
library(e1071)

# Linear kernel
svc.fit <- svm(default ~ balance + income, data = train_data, 
               kernel = "linear", cost = 10, scale = TRUE)

# Cross-validation to find best cost
tune.out <- tune(svm, default ~ balance + income, data = train_data, 
                 kernel = "linear", ranges = list(cost = c(0.1, 1, 10)))
best.svc <- tune.out$best.model
```

### Support Vector Machines (SVM)
Uses kernels (Radial, Polynomial) to create non-linear decision boundaries.
```{r}
# Radial Kernel
svm.radial <- svm(default ~ balance + income, data = train_data, 
                  kernel = "radial", gamma = 1, cost = 1)

# Plotting the boundary (Note: only works for 2 predictors)
plot(svm.radial, train_data, balance ~ income)

# Predictions
svm.pred <- predict(svm.radial, test_data)
table(predict = svm.pred, truth = test_data$default)
```


