---
title: "modeling-reference-sheet"
author: "Austin Melendez"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# STAT140B - Statistical Modeling
## Modeling Reference Cheat Sheet
### Austin Melendez
#### 3/5/2026

```{r}
#install.packages("MASS")
#install.packages("e1071")
#install.packages("class")
#install.packages("boot")
#install.packages("leaps")
#install.packages("glmnet")
#install.packages("pls")
```


# Chapter 2 - Simple Linear Regression
* Predict quantitative Y using one predictor

```{r}
# Fit model
fit <- lm(Y ~ X, data = mydata)

# Summary
summary(fit)

# Confidence intervals
confint(fit)

# Predictions
predict(fit, newdata=data.frame(X=5), interval="confidence")
predict(fit, newdata=data.frame(X=5), interval="prediction")
```

# Chapter 3 - Multiple Linear Regression
* Predict quantitative outputs using many predictors
* Check variable importance
* Can evaluate interactions and nonlinear terms
```{r}
fit <- lm(Y ~ X1 + X2 + X3, data = mydata)
summary(fit)

# Interaction
fit.int <- lm(Y ~ X1 * X2, data=mydata)

# Polynomial
fit.poly <- lm(Y ~ X1 + I(X1^2) + I(X1^3), data=mydata)

# Diagnostics
par(mfrow=c(2,2))
plot(fit)
```

# Chapter 4 - Classification Methods
### Logistic Regression
* Binary classifications
* Produces probabilities of landing in a class
* Interpretabble coefficients (log odds)
```{r}
log.fit <- glm(Y ~ X1 + X2, data=mydata, family=binomial)
summary(log.fit)

probs <- predict(log.fit, type="response")
preds <- ifelse(probs > 0.5, 1, 0)

mean(preds == mydata$Y)=
```

### Linear Discriminant Analysis (LDA)
* Classification when classes are well seperated
* Assumes predicators are normally dist
* Works better than logistic regression when $n$ is small
* Works when deicision boundaries are linear
```{r}
library(MASS)
lda.fit <- lda(Y ~ X1 + X2, data=mydata)

pred <- predict(lda.fit)$class

mean(pred == mydata$Y)
```

### Quadratic Discriminant Analysis (QDA)
* Like LDA but allows each class its own covariance
* Flexible, works when decision boundaries are nonlinear
```{r}
qda.fit <- qda(Y ~ X1 + X2, data=mydata)

pred <- predict(qda.fit)$class

mean(pred == mydata$Y)
```

### Naive Bayes
* Classifier assuming independece of predictors
* Works well with many predictors
* Good for text data
```{r}
library(e1071)

nb.fit <- naiveBayes(Y ~ X1 + X2, data=mydata)

pred <- predict(nb.fit, newdata=mydata)

mean(pred == mydata$Y)
```

### K-Nearest Neighbors (KNN)
* Nonparametric classification
* Good when decision boundary is irregular
* Must scale/normalize data!
```{r}
library(class)

train.X <- scale(train[, c("X1","X2")])
test.X  <- scale(test[, c("X1","X2")])

knn.pred <- knn(train.X, test.X, train$Y, k = 5)

mean(knn.pred == test$Y)
```

# Chapter 5 - Resampling Methods
### Validation Set Approach
* Split into train and test set
* Only trains on 50-70% of data (high variance)
* The model is trained on the training set and evaluated on the validation set.
```{r}
set.seed(1)
train.idx <- sample(1:nrow(mydata), nrow(mydata)/2)

train <- mydata[train.idx,]
test  <- mydata[-train.idx,]

fit <- lm(Y ~ ., data=train)

pred <- predict(fit, newdata=test)
mean((test$Y - pred)^2)
```

### Leave-One-Out Cross Validation (LOOCV)
* A special case of k-fold cross-validation where $k = n$
* Each observation is left out one at a time. Train model on $n−1$ observations, then test on the one left out observation. Average the prediction errors
* Use with small data sets, low bias and possible high variance
```{r}
library(boot)

loocv.err <- cv.glm(mydata, glm(Y ~ X1 + X2, data=mydata))

loocv.err$delta
```

### k-Fold Cross Validation
* The data is divided into $k$ equal-sized folds
* Split data into $k$ groups and train model on $k-1$ folds. Test on remaining folds and repeat until each fold has been used as the validation set. Then average the errors.
* Typical $k = 5$ or $k = 10$
* Use with moderate data sets, good balance between bias and variance
```{r}
set.seed(1)

cv.err <- cv.glm(mydata, glm(Y ~ X1 + X2, data=mydata), K=10)

cv.err$delta
```

### Bootstrap
* Bootstrap repeatedly samples observations with replacement from the dataset to create many simulated datasets called bootstrap samples
* Sample n observations with replacement, then fit model and compute statistic of interest (mean, median, CI, errors, bias, etc). Repeat process many times.
```{r}
library(boot)

boot.fn <- function(data, index) {
  coef(lm(medv ~ lstat + age, data=data, subset=index))
}

boot(Boston, boot.fn, R=1000)
```

# Chapter 6 - Model Selection and Regularization
## Select Subset of Predictors Methods
### Bet Subset Selection
* Best subset selection fits all possible models using every combination of predictors and selects the best model for each model size
* Use when the number of predictors p is small
* Uses Cp, BIC, adj-$R^2$
```{r}
library(leaps)

regfit <- regsubsets(Y ~ ., data=mydata, nvmax=10)
summary(regfit)

# Remove NA values
#Hitters <- na.omit(Hitters)

plot(summary(regfit)$bic, type="l")
plot(summary(regfit)$cp, type="l")
plot(summary(regfit)$adjr2, type="l")

which.min(regfit$cp)
which.min(regfit$bic)
which.max(regfit$adjr2)
```

### Forward / Backward Stepwise Selection
* Forward selection starts with no predictors and adds predictors one at a time
* Use forward selection when p is moderately large and works even when $p > n$
* Backward selection starts with the full model containing all predictors and removes the least useful predictor at each step
* Use backward selection when p is moderate and works only when $n > p$
* Both may miss the optimal model, depends on the order variables are added
```{r}
reg.fwd <- regsubsets(Y ~ ., data=mydata, method="forward")

reg.bwd <- regsubsets(Y ~ ., data=mydata, method="backward")
```

## Shrink Coefficient Methods
### Ridge Regression
* Ridge regression modifies least squares by adding a penalty on the squared size of coefficients
* Use with many predictors and possible multicollinearity
* You want coefficients to shrink towards zero but not be removed
* Use when many predictors exist and they are highly correlated. We can reduce variance without removing predictors
* Reduces overfitting and handles multicollinearity well
```{r}
library(glmnet)

X <- model.matrix(Y ~ ., data=mydata)[, -1]
y <- mydata$Y

ridge.mod <- glmnet(X, y, alpha=0)
cv.ridge  <- cv.glmnet(X, y, alpha=0)

best.lambda <- cv.ridge$lambda.min
coef(cv.ridge, s=best.lambda)
```

### Lasso
* Lasso adds a penalty based on the absolute value of coefficients, this penalty can shrink some coefficients exactly to zero
* Use when you believe only a subset of predictors are important
* Can behave unpredictably when predictors are highly correlated
```{r}
lasso.mod <- glmnet(X, y, alpha=1)
cv.lasso  <- cv.glmnet(X, y, alpha=1)

best.lambda <- cv.lasso$lambda.min
coef(cv.lasso, s=best.lambda)
```

## Dimension Reduction Methods
### Principal Components Regression (PCR)
* PCR first performs principal component analysis (PCA) on the predictors and then performs regression using the principal components instead of the original variables
* The components capture the directions of greatest variance in the predictors
* Use PCR when predictors are highly correlated, p is large, and we want to reduce dimensionality before regression
```{r}
library(pls)

pcr.fit <- pcr(Y ~ ., data=mydata, scale=TRUE, validation="CV")

summary(pcr.fit)
```

### Partial Least Squares (PLS)
* PLS is similar to PCR but differs because it chooses components using both predictors and the response variable
* This means the components are selected to maximize covariance with the response
* Use PLS when the predictors are highly correlated, you want dimension reduction, and you want components that directly help predict Y
```{r}
pls.fit <- plsr(Y ~ ., data=mydata, scale=TRUE, validation="CV")

summary(pls.fit)
```

