---
title: "lab-ch10"
author: "Austin Melendez"
date: "2026-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
#install.packages("torch")
#install.packages("luz")
#install.packages("ISLR2")
#install.packages("glmnet")
#install.packages("dplyr")


library(ISLR2)
library(torch)
library(luz)
library(glmnet)
library(dplyr)

```
Section 10.10: Exercises 7, 9

## Exercise 7
Fit a neural network to the Default data. Use a single hidden layer with 10 units, and dropout regularization. Have a look at Labs 10.9.1– 10.9.2 for guidance. Compare the classification performance of your model with that of linear logistic regression.

```{r}
set.seed(1)
torch_manual_seed(1)

# Load data
Default <- ISLR2::Default

# Convert response to numeric
Default$default_num <- ifelse(Default$default == "Yes", 1, 0)

# Create model matrix
x <- model.matrix(default ~ . - default_num - 1, data = Default) %>% scale()
y <- Default$default_num

# Train/test split
n <- nrow(Default)
testid <- sample(1:n, n / 3)
```

Lets do a Logistic Regression first:
```{r}
glm.fit <- glm(default_num ~ balance + income + student,
               data = Default[-testid, ],
               family = binomial)

glm.prob <- predict(glm.fit,
                    newdata = Default[testid, ],
                    type = "response")

glm.pred <- ifelse(glm.prob > 0.5, 1, 0)

glm.acc <- mean(glm.pred == y[testid])

print("Logistic Regression Accuracy:")
glm.acc
```

Then try a Neural Network from your reworked Chapter 10 Lab:
```{r}
modnn <- nn_module(
  initialize = function(input_size) {
    self$hidden <- nn_linear(input_size, 10)
    self$activation <- nn_relu()
    self$dropout <- nn_dropout(0.4)
    self$output <- nn_linear(10, 1)
  },

  forward = function(x) {
    x %>%
      self$hidden() %>%
      self$activation() %>%
      self$dropout() %>%
      self$output() %>%
      torch_flatten(start_dim = 1)
  }
)

modnn <- modnn %>%
  setup(
    loss = nn_bce_with_logits_loss(),
    optimizer = optim_rmsprop,
    metrics = list(luz_metric_binary_accuracy_with_logits())
  ) %>%
  set_hparams(input_size = ncol(x))

# Fit neural network
fitted <- modnn %>%
  fit(
    data = list(
      x[-testid, ],
      y[-testid]
    ),
    valid_data = list(
      x[testid, ],
      y[testid]
    ),
    epochs = 20
  )
```

```{r}
# Plot training history
plot(fitted)

# Neural network predictions
nn.prob <- predict(fitted, x[testid, ]) %>%
  torch_sigmoid() %>%
  as_array()

nn.pred <- ifelse(nn.prob > 0.5, 1, 0)

nn.acc <- mean(nn.pred == y[testid])

print("Neural Network Accuracy:")
nn.acc
```

```{r}
# Compare results
data.frame(
  Model = c("Logistic Regression", "Neural Network"),
  Accuracy = c(glm.acc, nn.acc)
)
```

We see that they actually perform very similarly, the Neural Network improved on the Logistic Regression's Accuracy by only $0.0015$ or $0.15$%.

## Exercise 9
Fit a lag-5 autoregressive model to the NYSE data, as described in the text and Lab 10.9.6. Refit the model with a 12-level factor representing the month. Does this factor improve the performance of the model?

```{r}
data("NYSE")

# Create lag variables
NYSE$lag1 <- c(NA, NYSE$DJ_return[-nrow(NYSE)])
NYSE$lag2 <- c(NA, NYSE$lag1[-nrow(NYSE)])
NYSE$lag3 <- c(NA, NYSE$lag2[-nrow(NYSE)])
NYSE$lag4 <- c(NA, NYSE$lag3[-nrow(NYSE)])
NYSE$lag5 <- c(NA, NYSE$lag4[-nrow(NYSE)])

# Remove missing rows
NYSE <- na.omit(NYSE)

# 12-level factor for the month
NYSE$date <- as.Date(NYSE$date)
NYSE$Month <- as.factor(format(NYSE$date, "%m"))
```

```{r}
n <- nrow(NYSE)

train <- 1:(n / 2)
test <- (n / 2 + 1):n
```

### Lag-5 Autoregressive Model
```{r}
fit1 <- lm(
  DJ_return ~ lag1 + lag2 + lag3 + lag4 + lag5,
  data = NYSE,
  subset = train
)

pred1 <- predict(fit1, newdata = NYSE[test, ])

mse1 <- mean(
  (NYSE$DJ_return[test] - pred1)^2
)


print("Lag-5 Model Test MSE:")
print(mse1)
```

### Lag-5 Model w/ 12-Month Factor
```{r}
fit2 <- lm(
  DJ_return ~ lag1 + lag2 + lag3 +
    lag4 + lag5 + Month,
  data = NYSE,
  subset = train
)

pred2 <- predict(fit2, newdata = NYSE[test, ])

mse2 <- mean(
  (NYSE$DJ_return[test] - pred2)^2
)

print("Lag-5 + Month Factor Test MSE:")
print(mse2)
```

Let's compare the two models:
```{r}
results <- data.frame(
  Model = c("Lag-5 AR", "Lag-5 AR + Month"),
  Test_MSE = c(mse1, mse2)
)

print(results)
```

Adding the month as a factor very slightly improved our model. It lowered the Test MSE by $0.005$.
