---
title: "lab-ch6-1"
author: "Austin Melendez"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("leaps")
#install.packages("glmnet")
library(leaps)
library(glmnet)
```


Section 6.6: Exercises 8, 10

## Exercise 8
### A
Generate a predictor X of length n = 100, as well as a noise vector $\epsilon$ of length n = 100.
```{r}
set.seed(1)
n <- 100
X <- rnorm(n)
eps <- rnorm(n, sd = 1)
```

### B
Generate a response vector Y of length n = 100.
```{r}
beta0 <- 1
beta1 <- 2
beta2 <- 3
beta3 <- 4

Y <- beta0 + beta1*X + beta2*X^2 + beta3*X^3 + eps
```

### C
```{r}
data <- data.frame(Y = Y, X = X)
for (i in 2:10) data[paste0("X", i)] <- X^i

best.fit <- regsubsets(Y ~ ., data = data, nvmax = 10)
best.sum <- summary(best.fit)
```

Let's plot $Cp$, $BIC$, and adjusted $R^2$.
```{r}
plot(best.sum$cp, type="b", main="Cp")
plot(best.sum$bic, type="b", main="BIC")
plot(best.sum$adjr2, type="b", main="Adjusted R2")
```

And then select the best model.
```{r}
which.min(best.sum$cp)
which.min(best.sum$bic)
which.max(best.sum$adjr2)
```

Let's extract the coefficients from our best model which uses 4 variables.
```{r}
coef(best.fit, which.min(best.sum$cp))
```

### D
Forward and Backward Stepwise Selection

```{r}
fwd.fit <- regsubsets(Y ~ ., data=data, nvmax=10, method="forward")

fwd.sum <- summary(fwd.fit)

which.min(fwd.sum$cp)
which.min(fwd.sum$bic)
which.max(fwd.sum$adjr2)
```
Let's extract the coefficients from our best models which again uses 4 variables.
```{r}
coef(best.fit, which.min(fwd.sum$cp))
```



```{r}
bwd.fit <- regsubsets(Y ~ ., data=data, nvmax=10, method="backward")
bwd.sum <- summary(bwd.fit)

which.min(bwd.sum$cp)
which.min(bwd.sum$bic)
which.max(bwd.sum$adjr2)
```
Let's extract the coefficients from our best models which again uses 4 variables.

```{r}
coef(best.fit, which.min(bwd.sum$cp))
```

### E
Lasso Model with Cross-Validation
```{r}
X.mat <- model.matrix(Y ~ ., data)[, -1]

set.seed(1)
cv.lasso <- cv.glmnet(X.mat, Y, alpha = 1)
plot(cv.lasso)
```

Let's find the best $\lambda$ and coefficients.
```{r}
best.lambda <- cv.lasso$lambda.min
best.lambda

coef(cv.lasso, s = best.lambda)
```

## Exercise 10
Generate data
### A
```{r}
set.seed(1)

n  <- 1000
p  <- 20

beta.true <- c(3, -2, 0, 0, 1.5, rep(0, p-5))

# X
X  <- matrix(rnorm(n * p), n, p)
eps <- rnorm(n)

# Response
Y <- X %*% beta.true + eps
```

### B
Train/Test split
```{r}
train.idx <- sample(1:n, 100)
test.idx  <- setdiff(1:n, train.idx)

X.train <- X[train.idx, ]
X.test  <- X[test.idx, ]
Y.train <- Y[train.idx]
Y.test  <- Y[test.idx]

mse <- function(y, yhat) mean((y - yhat)^2)
```

### C
Best subset selection
```{r}
train.data <- data.frame(Y = Y.train, X.train)

fit.best <- regsubsets(Y ~ ., data = train.data, nvmax = p)

# Extract MSE for each model size
train.mse <- rep(NA, p)

for (r in 1:p) {
  coef.r <- coef(fit.best, id = r)
  X.mat  <- model.matrix(Y ~ ., data = train.data)[, names(coef.r)]
  yhat   <- X.mat %*% coef.r
  train.mse[r] <- mse(Y.train, yhat)
}

plot(1:p, train.mse, type="l", lwd=2,
     xlab="Model Size", ylab="Training MSE",
     main="Training MSE vs Model Size")
```

### D
Test MSE for best model of each size
```{r}
test.data <- data.frame(Y = Y.test, X.test)
test.mse <- rep(NA, p)

for (r in 1:p) {
  coef.r <- coef(fit.best, id = r)
  X.mat  <- model.matrix(Y ~ ., data = test.data)[, names(coef.r)]
  yhat   <- X.mat %*% coef.r
  test.mse[r] <- mse(Y.test, yhat)
}

plot(1:p, test.mse, type="l", lwd=2,
     xlab="Model Size", ylab="Test MSE",
     main="Test MSE vs Model Size")
```

### E
Find model size that minimizes MSE
```{r}
best.size <- which.min(test.mse)
best.size
test.mse[best.size]
```

### F
Compare estimated coefficients and true $\beta$ value
```{r}
coef.best <- coef(fit.best, id = best.size)
coef.best
beta.true
```
