---
title: "lab5"
author: "Austin Melendez"
date: "2026-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Section 4.8: Exercises 5, 9

## Exercise 5
### A
```{r}
library(ISLR2)
data(Default)
<<<<<<< HEAD:STAT140B - Statistical Learning/Labs/lab5.Rmd


set.seed(1)

# Fit logistic regression using all data
log_mod <- glm(default ~ income + balance, data = Default, family = binomial)

summary(log_mod)
=======
set.seed(0)

log.mod <- lm(default ~ income + balance, data = Default)
>>>>>>> 3abdeeb80d2def0948581954ab4c64c0c8ed6f6e:STAT140B - Statistical Learning/Labs/lab-ch5.Rmd
```

### B
Validation Set Approach
```{r}
<<<<<<< HEAD:STAT140B - Statistical Learning/Labs/lab5.Rmd
# 50% training split
train.idx <- sample(1:nrow(Default), nrow(Default)/2)

train <- Default[train.idx, ]
valid <- Default[-train.idx, ]

# Fit logistic regression on training set
log_train <- glm(default ~ income + balance, data=train, family=binomial)

prob <- predict(log_train, newdata=valid, type="response")
pred <- ifelse(prob > 0.5, "Yes", "No")

# Calculate error rate using validation set
val.error1 <- mean(pred != valid$default)
val.error1
=======
train <- sample(10000, 7000)
>>>>>>> 3abdeeb80d2def0948581954ab4c64c0c8ed6f6e:STAT140B - Statistical Learning/Labs/lab-ch5.Rmd
```

### C
Repeat three times with different random splits
```{r}
errors <- rep(0, 3)

for (i in 1:3) {
  train.idx <- sample(1:nrow(Default), nrow(Default)/2)
  
  train <- Default[train.idx, ]
  valid <- Default[-train.idx, ]
  
  log_train2 <- glm(default ~ income + balance, data=train, family=binomial)
  
  prob <- predict(log_train2, newdata=valid, type="response")
  pred <- ifelse(prob > 0.5, "Yes", "No")
  
  errors[i] <- mean(pred != valid$default)
}

errors
mean(errors)
```
The validation error for each subset varies slightly between the different splits. Each split got a validation error rate of $0.0274$, $0.0244$, and $0.0244$ with a average error rate of $0.0254$ or $2.54\%$.


### D
Add student dummy variable
```{r}
train.idx <- sample(1:nrow(Default), nrow(Default)/2)

train <- Default[train.idx, ]
valid <- Default[-train.idx, ]

log_student <- glm(default ~ income + balance + student,
                   data=train, family=binomial)

prob2 <- predict(log_student, newdata=valid, type="response")
pred2 <- ifelse(prob2 > 0.5, "Yes", "No")

val.error.student <- mean(pred2 != valid$default)
val.error.student
```
Adding the categorical variable `student` does not improve predictive performance of the model even though the coefficient might be statistically significant. Our validation error rate using this new model is worse, being at $0.0278$.

## Exercise 9
### A
Estimate the population mean of `medv`
```{r}
data(Boston)
<<<<<<< HEAD:STAT140B - Statistical Learning/Labs/lab5.Rmd

mu_hat <- mean(Boston$medv)
mu_hat
=======
train2 <- sample(506, 355)

lm2 <- lm(mpg ~ horsepower, data = Boston, subset = train)
>>>>>>> 3abdeeb80d2def0948581954ab4c64c0c8ed6f6e:STAT140B - Statistical Learning/Labs/lab-ch5.Rmd
```
This is the sample mean, our estimate of the true population average of median home value.

### B
Estimate the standard error of `mu hat`
```{r}
se_mean <- sd(Boston$medv) / sqrt(nrow(Boston))
se_mean
```
This is how much our estimate of the mean would vary if we repeatedly sampled new data sets of the same size.

### C
Bootstrap estimate of the SE of the mean
```{r}
B <- 1000
boot.means <- rep(0, B)

for (b in 1:B) {
  boot.sample <- sample(Boston$medv, replace = TRUE)
  boot.means[b] <- mean(boot.sample)
}

boot_se.mean <- sd(boot.means)
boot_se.mean
```
Our bootstrap SE is very close to our estimate from B

### D
95% CI using bootstrap SE vs t test
```{r}
ci_boot <- c(mu_hat - 2*boot_se.mean, mu_hat + 2*boot_se.mean)
ci_boot
```


### E
Estimate the population median
```{r}
median_hat <- median(Boston$medv)
median_hat
```

### F
Bootstrap SE of the median
```{r}
boot.medians <- rep(0, B)

for (b in 1:B) {
  boot.sample <- sample(Boston$medv, replace = TRUE)
  boot.medians[b] <- median(boot.sample)
}

boot.se.median <- sd(boot.medians)
boot.se.median
```
SE of the median is larger than SE of the mean because the median is more variable.

### G
Estimate the 10th percentile of `medv`
```{r}
p10_hat <- quantile(Boston$medv, 0.10)
p10_hat
```

### H
Bootstrap SE of the 10th percentile
```{r}
boot.p10 <- rep(0, B)

for (b in 1:B) {
  boot.sample <- sample(Boston$medv, replace = TRUE)
  boot.p10[b] <- quantile(boot.sample, 0.10)
}

boot.se.p10 <- sd(boot.p10)
boot.se.p10
```
Percentiles near the tails often have higher SE than central measures. (like the mean)
