---
title: "lab-ch6-2"
author: "Austin Melendez"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Section 6.6: Exercises 9, 11

```{r}
library(ISLR2)
library(glmnet)
library(pls)
library(leaps)
```

## Exercise 9
Predict the number of applications received using the other variables in the `College` data set.

### A
Split the data into Train/Test sets

```{r}
data(College)

set.seed(1)
train <- sample(1:nrow(College), nrow(College)/2)

test <- -train
```

### B
Fit a linear model using least squares on the training set, and report the test error.

```{r}
lm.fit <- lm(Apps ~ ., data=College, subset=train)

lm.pred <- predict(lm.fit, College[test,])

lm.test.mse <- mean((College$Apps[test] - lm.pred)^2)

lm.test.mse
```

### C
Fit a ridge regression model on the training set, with $\lambda$ chosen by cross-validation. Report the test error.
```{r}
x <- model.matrix(Apps~., College)[,-1]
y <- College$Apps

set.seed(1)

cv.ridge <- cv.glmnet(x[train,], y[train], alpha=0)

ridge.pred <- predict(cv.ridge,
                      s=cv.ridge$lambda.min,
                      newx=x[test,])

ridge.mse <- mean((ridge.pred - y[test])^2)

ridge.mse
```

### D
Fit a lasso model on the training set, with $\lambda$ chosen by cross-validation. Report the test error and number of non-zero coefficient estimates.
```{r}
set.seed(1)

cv.lasso <- cv.glmnet(x[train,], y[train], alpha=1)

lasso.pred <- predict(cv.lasso,
                      s=cv.lasso$lambda.min,
                      newx=x[test,])

lasso.mse <- mean((lasso.pred - y[test])^2)

lasso.mse
```

Number of non-zero coefficients:
```{r}
lasso.coef <- predict(cv.lasso,
                      type="coefficients",
                      s=cv.lasso$lambda.min)

sum(lasso.coef != 0)
```

### E
Fit a PCR model on the training set, with $M$ chosen by cross-validation. Report the test error obtained, along with the value of $M$ selected by cross-validation.
```{r}
set.seed(1)

pcr.fit <- pcr(Apps~., data=College,
               subset=train,
               scale=TRUE,
               validation="CV")

validationplot(pcr.fit, val.type="MSEP")

pcr.pred <- predict(pcr.fit,
                    College[test,],
                    ncomp=10)

pcr.mse <- mean((pcr.pred - College$Apps[test])^2)

pcr.mse
```

$M$ value:
```{r}
summary(pcr.fit)
```


### F
Fit a PLS model on the training set, with $M$ chosen by cross-validation. Report the test error obtained, along with the value of $M$ selected by cross-validation.
```{r}
set.seed(1)

pls.fit <- plsr(Apps~., data=College,
                subset=train,
                scale=TRUE,
                validation="CV")

validationplot(pls.fit, val.type="MSEP")

pls.pred <- predict(pls.fit,
                    College[test,],
                    ncomp=10)

pls.mse <- mean((pls.pred - College$Apps[test])^2)

pls.mse
```

$M$ value:
```{r}
summary(pls.fit)
```

### G
How accurately can we predict the number of college applications received? Is there much difference among the test errors resulting from these five approaches?
```{r}
data.frame(
Method=c("Linear","Ridge","Lasso","PCR","PLS"),
Test_MSE=c(lm.test.mse,ridge.mse,lasso.mse,pcr.mse,pls.mse))
```


## Exercise 11
We will now try to predict per capita crime rate in the Boston data set.
### A
#### Best Subset
```{r}
data(Boston)

set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston)/2)
test <- -train

regfit <- regsubsets(crim ~ ., data = Boston[train,], nvmax = 13)

reg.summary <- summary(regfit)

predict.regsubsets <- function(object, newdata, id){
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form, newdata)
  coefi <- coef(object, id = id)
  xvars <- names(coefi)
  mat[, xvars] %*% coefi
}

n.models <- length(reg.summary$rsq)

val.errors <- rep(NA, n.models)

for(i in 1:n.models){
  pred <- predict.regsubsets(regfit, Boston[test,], id = i)
  
  val.errors[i] <- mean((Boston$crim[test] - pred)^2)
}

val.errors[which.min(val.errors)]
```

#### Ridge Regression
```{r}
x <- model.matrix(crim~., Boston)[,-1]
y <- Boston$crim

cv.ridge <- cv.glmnet(x[train,],y[train],alpha=0)

ridge.pred <- predict(cv.ridge,
                      s=cv.ridge$lambda.min,
                      newx=x[test,])

mean((ridge.pred-y[test])^2)
```

#### Lasso
```{r}
cv.lasso <- cv.glmnet(x[train,],y[train],alpha=1)

lasso.pred <- predict(cv.lasso,
                      s=cv.lasso$lambda.min,
                      newx=x[test,])

mean((lasso.pred-y[test])^2)
```

#### Principal Component Regression (PCR)
```{r}
pcr.fit <- pcr(crim~., data=Boston,
               subset=train,
               scale=TRUE,
               validation="CV")

pcr.pred <- predict(pcr.fit,
                    Boston[test,],
                    ncomp=6)

mean((pcr.pred-Boston$crim[test])^2)
```

### B
Our best models are produces using ridge regression, lasso, and best subset selection. These methods all balance bias and variance well, producing a strong interpretable result. I would choose the best subset selection model since it has the lowest validation set error of $40.14557$.

### C
No, the model does not use all predictors. This is because some predictors contribute little helpful information to our model. Reducing these less helpful predictors can help reduce and prevent the risk of overfitting. 