---
title: "lab-ch7"
author: "Austin Melendez"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Section 7.9: Exercises 8, 9, 11

```{r}
library(ISLR2)
library(glmnet)
library(pls)
library(leaps)
```

## Exercise 8
Fit some of the non-linear models investigated in this chapter to the `Auto` data set. Is there evidence for non-linear relationships in this data set?

### A
```{r}
library(ISLR2)
library(splines)

data(Auto)

# Polynomial
fit_poly <- lm(mpg ~ poly(horsepower, 4), data = Auto)

# Regression spline
fit_bs <- lm(mpg ~ bs(horsepower, df = 5), data = Auto)

summary(fit_poly)
summary(fit_bs)
```

```{r}
hp_grid <- seq(min(Auto$horsepower), max(Auto$horsepower), length = 100)

pred_poly <- predict(fit_poly, newdata = list(horsepower = hp_grid))
pred_bs <- predict(fit_bs, newdata = list(horsepower = hp_grid))

plot(Auto$horsepower, Auto$mpg, col = "grey",
     xlab = "Horsepower", ylab = "MPG")

lines(hp_grid, pred_poly, col = "blue", lwd = 2)
lines(hp_grid, pred_bs, col = "red", lwd = 2)

legend("topright", legend = c("Polynomial", "Spline"),
       col = c("blue","red"), lwd = 2)
```

* Strong evidence of non-linearity.
* `MPG` declines rapidly at low horsepower, then levels off.
* Splines capture this pattern more smoothly than polynomials.

## Exercise 9
From the `Boston` data, we will treat `dis` as the predictor and `nox` as the response.

### A
Use the poly() function to fit a cubic polynomial regression to predict `nox` using `dis`.
```{r}
data(Boston)

fit3 <- lm(nox ~ poly(dis, 3), data = Boston)
summary(fit3)

dis_grid <- seq(min(Boston$dis), max(Boston$dis), length = 100)
pred3 <- predict(fit3, newdata = list(dis = dis_grid))

plot(Boston$dis, Boston$nox, col = "grey")
lines(dis_grid, pred3, col = "blue", lwd = 2)
```

### B
Plot the polynomial fits for a range of different polynomial degrees (from 1 to 10).
```{r}
rss <- rep(NA, 10)

for (i in 1:10) {
  fit <- lm(nox ~ poly(dis, i), data = Boston)
  rss[i] <- sum(residuals(fit)^2)
}

plot(1:10, rss, type = "b", xlab = "Degree", ylab = "RSS")

print(mean(rss))
```

### C
Perform cross-validation or another approach to select the optimal degree for the polynomial.
```{r}
library(boot)

cv_error <- rep(NA, 10)

for (i in 1:10) {
  fit <- glm(nox ~ poly(dis, i), data = Boston)
  cv_error[i] <- cv.glm(Boston, fit, K = 10)$delta[1]
}

plot(1:10, cv_error, type = "b",
     xlab = "Degree", ylab = "CV Error")
```
I would select degree 3 or 4 since it has the lowest CV error rate.

### D
Use the bs() function to fit a regression spline to predict `nox` using `dis`. Report the output for the fit using four degrees of freedom.
```{r}
fit_bs <- lm(nox ~ bs(dis, df = 4), data = Boston)
summary(fit_bs)

pred_bs <- predict(fit_bs, newdata = list(dis = dis_grid))

plot(Boston$dis, Boston$nox, col = "grey")
lines(dis_grid, pred_bs, col = "red", lwd = 2)
```

### E
Now fit a regression spline for a range of degrees of freedom, and plot the resulting fits and report the resulting RSS.
```{r}
rss_spline <- rep(NA, 10)

for (i in 3:10) {
  fit <- lm(nox ~ bs(dis, df = i), data = Boston)
  rss_spline[i] <- sum(residuals(fit)^2)
}

plot(3:10, rss_spline[3:10], type = "b",
     xlab = "DF", ylab = "RSS")
```
RSS is lowest when degrees of freedom is equal to 10. More generally, RSS decreases with df.

### F
Perform cross-validation or another approach in order to select the best degrees of freedom for a regression spline on this data.
```{r}
cv_spline <- rep(NA, 10)

for (i in 3:10) {
  fit <- glm(nox ~ bs(dis, df = i), data = Boston)
  cv_spline[i] <- cv.glm(Boston, fit, K = 10)$delta[1]
}

plot(3:10, cv_spline[3:10], type = "b",
     xlab = "DF", ylab = "CV Error")
```
However, using CV we see that the optimal df is actually moderate (about 7). This confirms bias-variance tradeoff.

## Exercise 11
In Section 7.7, it was mentioned that GAMs are generally fit using a backfitting approach. We will now explore backfitting in the context of multiple linear regression.

### A
Generate a response $Y$ and two predictors $X_1$ and $X_2$, with $n = 100$.
```{r}
set.seed(1)
n <- 100

x1 <- rnorm(n)
x2 <- rnorm(n)
y <- 3 + 2*x1 + 4*x2 + rnorm(n)
```

### B-E
Backfitting loop.
```{r}
beta1 <- 100
beta2 <- -100

beta0_vals <- beta1_vals <- beta2_vals <- rep(NA, 1000)

for (i in 1:1000) {

  # update beta2
  a <- y - beta1 * x1
  beta2 <- lm(a ~ x2)$coef[2]

  # update beta1
  a <- y - beta2 * x2
  beta1 <- lm(a ~ x1)$coef[2]

  # update beta0
  beta0 <- mean(y - beta1*x1 - beta2*x2)

  beta0_vals[i] <- beta0
  beta1_vals[i] <- beta1
  beta2_vals[i] <- beta2
}

plot(beta1_vals, type = "l", col = "blue",
     ylim = range(c(beta0_vals, beta1_vals, beta2_vals)),
     ylab = "Coefficient Value",
     xlab = "Iteration")
lines(beta2_vals, col = "red")
lines(beta0_vals, col = "green")

legend("topright", legend = c("beta1","beta2","beta0"),
       col = c("blue","red","green"), lwd = 2)

```
This plot looks wrong... unless the coefficient values are being converged too very rapidly? 

### F
Compare your answer in (e) to the results of simply performing multiple linear regression to predict $Y$ using $X_1$ and $X_2$.
```{r}
fit <- lm(y ~ x1 + x2)


```

### G
On this data set, how many backfitting iterations were required in order to obtain a “good” approximation to the multiple regression coefficient estimates?

* Converges quickly, almost instantly
* Final estimates approx equal multiple regression estimates
