---
title: "lab-ch8"
author: "Austin Melendez"
date: "2026-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Section 8.4: Exercises 8, 9

## Exercise 8
In the lab, a classification tree was applied to the `Carseats` data set after converting `Sales` into a qualitative response variable. Now we will seek to predict `Sales` using regression trees and related approaches, treating the response as a quantitative variable.

```{r}
#install.packages("ISLR2")
#install.packages("tree")
#install.packages("randomForest")
#install.packages("BART")

library(ISLR2)
library(tree)

set.seed(1)

data(Carseats)
```

### A
Split into training and test set.
```{r}
train <- sample(1:nrow(Carseats), nrow(Carseats)/2)
Carseats_train <- Carseats[train, ]
Carseats_test  <- Carseats[-train, ]
```

### B
Fit a regression tree to the training set. Plot the tree and interpret the results.
```{r}
tree_fit <- tree(Sales ~ ., data = Carseats_train)

# Plot tree
plot(tree_fit)
text(tree_fit, pretty = 0)

tree_pred <- predict(tree_fit, newdata = Carseats_test)

# Test MSE
mean((tree_pred - Carseats_test$Sales)^2)
```

### C
Use cross-validation to determine the optimal level of tree complexity. Does pruning the tree improve the test MSE?
```{r}
set.seed(1)

cv_tree <- cv.tree(tree_fit)

# Plot CV results
plot(cv_tree$size, cv_tree$dev, type = "b",
     xlab = "Tree Size", ylab = "Deviance")

# Find best size using CV
best_size <- cv_tree$size[which.min(cv_tree$dev)]

# Prune tree
pruned_tree <- prune.tree(tree_fit, best = best_size)

# Plot pruned tree
plot(pruned_tree)
text(pruned_tree, pretty = 0)

# Test MSE
pruned_pred <- predict(pruned_tree, newdata = Carseats_test)
mean((pruned_pred - Carseats_test$Sales)^2)
```

### D
Use the bagging approach. What test MSE do you obtain? Use `importance()` to determine which variables are most important.
```{r}
library(randomForest)

set.seed(1)

bag_fit <- randomForest(Sales ~ ., data = Carseats_train,
                        mtry = ncol(Carseats_train) - 1,
                        importance = TRUE)

# Predictions
bag_pred <- predict(bag_fit, newdata = Carseats_test)

# Test MSE
mean((bag_pred - Carseats_test$Sales)^2)

# Variable importance
importance(bag_fit)
varImpPlot(bag_fit)
```

### E
Use random forests. What test MSE do you obtain? Use `importance()` to determine which variables are most important. Describe the effect of $m$, the number of variables onsidered at each split, and the error rate obtained.
```{r}
set.seed(1)

rf_fit <- randomForest(Sales ~ ., data = Carseats_train,
                       mtry = floor(sqrt(ncol(Carseats_train) - 1)),
                       importance = TRUE)

# Predictions
rf_pred <- predict(rf_fit, newdata = Carseats_test)

# Test MSE
mean((rf_pred - Carseats_test$Sales)^2)

# Importance
importance(rf_fit)
varImpPlot(rf_fit)
```

### F
Analyze the data using BART.
```{r}
library(BART)

set.seed(1)

# Prepare matrices
x_train <- model.matrix(Sales ~ . - 1, data = Carseats_train)
y_train <- Carseats_train$Sales
x_test  <- model.matrix(Sales ~ . - 1, data = Carseats_test)

# Fit BART
bart_fit <- gbart(x_train, y_train, x_test)

# Predictions
bart_pred <- bart_fit$yhat.test.mean

# Test MSE
mean((bart_pred - Carseats_test$Sales)^2)
```


## Exercise 9
This problem invovles the `OJ` data set which is part of the `ISLR2` package.
```{r}
library(ISLR2)
library(tree)

set.seed(1)

data(OJ)
```


### A
Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.
```{r}
train <- sample(1:nrow(OJ), 800)
OJ_train <- OJ[train, ]
OJ_test  <- OJ[-train, ]
```

### B
Fit a tree to the training data, with `Purchase` as the response and all others as predictors. What is the training error rate? How many terminal nodes does the tree have?
```{r}
oj_tree <- tree(Purchase ~ ., data = OJ_train)

summary(oj_tree)
```

### C
Pick one of the terminal nodes and interpret the information displayed after typing in the name of a tree object to get a detailed text output.
```{r}
oj_tree
```
To select a terminal node, we look for nodes that have a *, this idicates that it is a terminal node.

Node 7 corresponds to the observations where `LoyalCH` $> 0.764572$. This node contains $261$ observations and the predicted class for this node is `CH`.

It also shows us that about 95.8% of the observations purchased `CH` and 4.2% purchased `MM`.

This means that customers with a high value of `LoyalCH` are very liekly to purchase `CH`.

### D
Create a plot of the tree.
```{r}
plot(oj_tree)
text(oj_tree, pretty = 0)
```

### E
Predict the response on the test data, and produce a confusion matrix comparing the test labels and predicted test labels. What is the test error rate?
```{r}
oj_pred <- predict(oj_tree, newdata = OJ_test, type = "class")

table(oj_pred, OJ_test$Purchase)

# Test error rate
mean(oj_pred != OJ_test$Purchase)
```

### F
Apply the `cv.tree()` function to the training set in order to determine the optimal tree size.
```{r}
set.seed(1)

cv_oj <- cv.tree(oj_tree, FUN = prune.tree)
```

### G
Produce a plot with tree size on the x-axis and cross-validated classification error rate on the y-axis.
```{r}
plot(cv_oj$size, cv_oj$dev, type = "b",
     xlab = "Tree Size", ylab = "CV Error")
```

### H
Which tree size corresponds to the lowest cross-validated classification error rate?
```{r}
best_size <- cv_oj$size[which.min(cv_oj$dev)]
best_size
```

### I
Produce a pruned tree corresponding to the optimal tree size obtained using CV.
```{r}
pruned_oj <- prune.tree(oj_tree, best = best_size)

plot(pruned_oj)
text(pruned_oj, pretty = 0)
```

### J
Compare the training error rates between the pruned and unpruned trees. Which is higher?
```{r}
# Unpruned
train_pred_unpruned <- predict(oj_tree, type = "class")
mean(train_pred_unpruned != OJ_train$Purchase)

# Pruned
train_pred_pruned <- predict(pruned_oj, type = "class")
mean(train_pred_pruned != OJ_train$Purchase)
```
The pruned tree's training error is higher than the unpruned tree.

### K
Compare the test error rates between the pruned and unpruned trees. Which is higher?
```{r}
# Unpruned
test_pred_unpruned <- predict(oj_tree, newdata = OJ_test, type = "class")
mean(test_pred_unpruned != OJ_test$Purchase)

# Pruned
test_pred_pruned <- predict(pruned_oj, newdata = OJ_test, type = "class")
mean(test_pred_pruned != OJ_test$Purchase)
```
The pruned tree's test error rate is very slightly higher than the unpruned tree.
