---
title: "lab4-1"
author: "Austin Melendez"
date: "2026-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Section 4.8: Exercises 13 & 15

## Exercise 13

### A

```{r}
library(ISLR2)
library(ggplot2)
data("Weekly")

summary(Weekly)


plot(Weekly)
```
There seems to be a good spread of data across 20 years. Nothing looks too crazy.

### B
```{r}
log.fit <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = Weekly, family = binomial)

summary(log.fit)
```
The variable `Lag2` seems to be statistically significant at the 0.05 level.

### C
```{r}
log.prob <- predict(log.fit, type = "response")
log.pred <- ifelse(log.prob > 0.5, "Up", "Down")
table(Predicted = log.pred, Actual = Weekly$Direction)

mean(log.pred == Weekly$Direction)
```
This logistic regression has a tendency to over predict the `Up` category and rarely predict `Down`. It therefore frequently miss classifies the `Down` class as being `Up`.

### D
```{r}
train <- ( Weekly$Year <= 2008 )

log.fit2 <- glm(Direction ~ Lag2, data=Weekly, family=binomial, subset=train)
log.prob2 <- predict(log.fit2, newdata=Weekly[!train,], type="response")
log.pred2 <- ifelse(log.prob2 > 0.5, "Up", "Down")

table(Predicted = log.pred2, Actual = Weekly$Direction[!train])
mean(log.pred2 == Weekly$Direction[!train])
```
The model using train/test data split is more accurate, but it still over classifies the `Up` class. However the model accuracy has improved from $56.10%$ to $62.50%$.

### E
```{r}
library(MASS)
lda.fit <- lda(Direction ~ Lag2, data=Weekly, subset=train)
lda.pred <- predict(lda.fit, Weekly[!train,])
table(Predicted = lda.pred$class, Actual = Weekly$Direction[!train])
mean(lda.pred$class == Weekly$Direction[!train])
```
### F
```{r}
qda.fit <- qda(Direction ~ Lag2, data=Weekly, subset=train)
qda.pred <- predict(qda.fit, Weekly[!train,])
table(Predicted = qda.pred$class, Actual = Weekly$Direction[!train])
mean(qda.pred$class == Weekly$Direction[!train])
```
### G
```{r}
library(class)
train.X <- as.matrix(Weekly$Lag2[train])
test.X  <- as.matrix(Weekly$Lag2[!train])
train.Direction <- Weekly$Direction[train]

set.seed(1)
knn.pred <- knn(train.X, test.X, train.Direction, k = 1)

table(Predicted = knn.pred, Actual = Weekly$Direction[!train])
mean(knn.pred == Weekly$Direction[!train])
```
### H 
```{r}
library(e1071)
nb.fit <- naiveBayes(Direction ~ Lag2, data=Weekly, subset=train)
nb.pred <- predict(nb.fit, Weekly[!train,])
table(Predicted = nb.pred, Actual = Weekly$Direction[!train])
mean(nb.pred == Weekly$Direction[!train])
```
### I
| Method            | Accuracy |
|-------------------|----------|
| Logistic (Lag2)   | 62.50%   |
| LDA (Lag2)        | 62.50%   |
| QDA (Lag2)        | 58.65%   |
| KNN (K = 1)       | 50.00%   |
| Naive Bayes       | 58.65%   |

Based on my models above, the best model are the `LDA` and `Logistic` models.

### J
Let's try some transformations:
```{r}
Weekly$Lag2_sq <- Weekly$Lag2^2
Weekly$Lag1_Lag2 <- Weekly$Lag1 * Weekly$Lag2

mod <- glm(Direction ~ Lag2 + Lag2_sq, family=binomial,
           data=Weekly, subset=train)
pred <- ifelse(predict(mod, Weekly[!train,], type="response") > 0.5, "Up", "Down")
mean(pred == Weekly$Direction[!train])
```

Let's also try KNN with an optimized K value:
```{r}
acc <- c()
for (k in 1:20) {
  pred <- knn(train.X, test.X, train.Direction, k = k)
  acc[k] <- mean(pred == Weekly$Direction[!train])
}
which.max(acc) ; max(acc)
```
## Exercise 15

### A
```{r}
Power <- function() {
  print(2^3)
}
```

### B
```{r}
Power2 <- function(x, a) {
  print(x^a)
}
```

### C
```{r}
Power2(10, 3) # Prints 1000
Power2(8, 17) # Prints 2251799813685248
Power2(13, 13) # Prints 302875106592253
```

### D
```{r}
Power3 <- function(x, a) {
  result <- x^a
  return(result)
}

Power3(2, 5) # Prints 32
```

### E
```{r}
x <- 1:10
y <- Power3(x, 2)

plot(x, y,
     xlab = "x",
     ylab = "x^2",
     main = "Plot of f(x) = x^2 for x = 1 to 10")
```

### F
```{r}
PlotPower <- function(x, a) {
  y <- x^a
  plot(x, y,
       xlab = "x",
       ylab = paste("x^", a, sep=""),
       main = paste("Plot of f(x) = x^", a, sep=""))
}

PlotPower(1:10, 3)
```


