---
title: "lab-ch10"
author: "Austin Melendez"
date: "2026-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
#install.packages("keras")
#install.packages("magick")
#install.packages("ISLR2")
#install.packages("torch")
#install.packages("torchvision")

library(keras)
library(magick)
library(ISLR2)
library(torch)
library(torchvision)
```
Section 10.10: Exercises 8, 9

## Exercise 8
From your collection of personal photographs, pick 10 images of animals (such as dogs, cats, birds, farm animals, etc.). If the subject does not occupy a reasonable part of the image, then crop the image.

Now use a pretrained image classification CNN as in Lab 10.9.4 to predict the class of each of your images, and report the probabilities for the top five predicted classes for each image.

### (From Your Revised Textbook Lab 10) 
### Using Pretrained CNN Models
```{r}
setwd("C:/Users/Austin/Desktop/Everything/Coding/sacramento-state-university/STAT140B - Statistical Learning/Labs/") # you will need to update this! #
img_dir <- "lab-ch10-data"
image_names <- list.files(img_dir)
num_images <- length(image_names)
x <- torch_empty(num_images, 3, 224, 224)
for (i in 1:num_images) {
   img_path <- file.path(img_dir, image_names[i])
   img <- img_path %>% 
     base_loader() %>% 
     transform_to_tensor() %>% 
     transform_resize(c(224, 224)) %>% 
     # normalize with imagenet mean and stds.
     transform_normalize(
       mean = c(0.485, 0.456, 0.406),
       std = c(0.229, 0.224, 0.225)
     )
   x[i,,, ] <- img
}
```

We then load the trained network. The model has 18 layers, with a fair bit of complexity.

```{r}
model <- torchvision::model_resnet18(pretrained = TRUE)
model$eval() # put the model in evaluation mode
```

Finally, we classify our images, and return the top five class
choices in terms of predicted probability for each.

```{r}
preds <- model(x)

mapping <- jsonlite::read_json("https://s3.amazonaws.com/deep-learning-models/image-models/imagenet_class_index.json") %>% 
  sapply(function(x) x[[2]])

top5 <- torch_topk(preds, dim = 2, k = 5)

top5_prob <- top5[[1]] %>% 
  nnf_softmax(dim = 2) %>% 
  torch_unbind() %>% 
  lapply(as.numeric)

top5_class <- top5[[2]] %>% 
  torch_unbind() %>% 
  lapply(function(x) mapping[as.integer(x)])

result <- purrr::map2(top5_prob, top5_class, function(pr, cl) {
  names(pr) <- cl
  pr
})
names(result) <- image_names
print(result)
```
For reference, the images I used are uploaded and labeled here on [Imgur.com](https://imgur.com/a/DraPV2s):

## Exercise 9
Fit a lag-5 autoregressive model to the NYSE data, as described in the text and Lab 10.9.6. Refit the model with a 12-level factor representing the month. Does this factor improve the performance of the model?

```{r}
data("NYSE")

# Create lag variables
NYSE$lag1 <- c(NA, NYSE$DJ_return[-nrow(NYSE)])
NYSE$lag2 <- c(NA, NYSE$lag1[-nrow(NYSE)])
NYSE$lag3 <- c(NA, NYSE$lag2[-nrow(NYSE)])
NYSE$lag4 <- c(NA, NYSE$lag3[-nrow(NYSE)])
NYSE$lag5 <- c(NA, NYSE$lag4[-nrow(NYSE)])

# Remove missing rows
NYSE <- na.omit(NYSE)

# 12-level factor for the month
NYSE$date <- as.Date(NYSE$date)
NYSE$Month <- as.factor(format(NYSE$date, "%m"))
```

```{r}
n <- nrow(NYSE)

train <- 1:(n / 2)
test <- (n / 2 + 1):n
```

### Lag-5 Autoregressive Model
```{r}
fit1 <- lm(
  DJ_return ~ lag1 + lag2 + lag3 + lag4 + lag5,
  data = NYSE,
  subset = train
)

pred1 <- predict(fit1, newdata = NYSE[test, ])

mse1 <- mean(
  (NYSE$DJ_return[test] - pred1)^2
)


print("Lag-5 Model Test MSE:")
print(mse1)
```

### Lag-5 Model w/ 12-Month Factor
```{r}
fit2 <- lm(
  DJ_return ~ lag1 + lag2 + lag3 +
    lag4 + lag5 + Month,
  data = NYSE,
  subset = train
)

pred2 <- predict(fit2, newdata = NYSE[test, ])

mse2 <- mean(
  (NYSE$DJ_return[test] - pred2)^2
)

print("Lag-5 + Month Factor Test MSE:")
print(mse2)
```

Let's compare the two models:
```{r}
results <- data.frame(
  Model = c("Lag-5 AR", "Lag-5 AR + Month"),
  Test_MSE = c(mse1, mse2)
)

print(results)
```

Adding the month as a factor very slightly improved our model. It lowered the Test MSE by $0.005$.
